```{r}
install.packages(c("dplyr", "ggplot2", "maps", "patchwork", "tidyverse", "FRK", "lubridate", 
                   "MASS", "gstat", "sp", "spacetime", "animation"))

```


## Our dataset: NOAA daily maximum temperature

We will use the **NOAA data set**: daily maximum temperature (`Tmax`) in degrees Celsius (°C) at **152 weather stations** in the central and eastern USA, recorded in **2023**. We will focus on **May – September 2023** only in our examples.

```{r}
NOAA_df <- read.csv("NOAA_2023 v2.csv", header=TRUE, sep=",") # read in the dataset
# Create the new "date" column
NOAA_df$date <- as.Date(paste(NOAA_df$year, NOAA_df$month, NOAA_df$day, sep = "-"))
head(NOAA_df)
```


```{r, warning=FALSE}
library(dplyr) # need to apply functions in the package dplyr and have to load it here
Tmax <- filter(NOAA_df,               # only retrieve the variables and range of time we need
               var  == "Tmax",        # daily maximum temperature
               month %in% 5:9,        # May–September
               year  == 2023)         # year 2023

Tmax$t <- as.numeric(Tmax$date) - as.numeric(as.Date("2023-04-30"))          # day 1 = May 1, 2023

```

# PART 1 — Visualizing Spatio-Temporal Data

A **spatial plot** shows the value of a variable at every location, for **one time point**.

Let's plot daily maximum temperature on three different days: **May 1, June 1, July 1 of 2023**.

```{r spatial-plot, fig.height=4}
library(ggplot2) # a useful package for statistical plots
Tmax_1 <- subset(Tmax, t %in% c(1, 32, 62)) # retrieve data on the 3 days only

NOAA_plot <- ggplot(Tmax_1) +  # Create a ggplot object using Tmax_1 as the main dataset.
  geom_point(aes(x = lon, y = lat, colour = z), size = 2) +  # Add points with longitude and latitude, 
  # colored by z, with size 2.
  scale_colour_viridis_c(name = "degC", option = "turbo") +  # Apply a continuous viridis color scale 
  # labeled "degC" using the turbo palette.
  geom_path(data = map_data("state"),
            aes(x = long, y = lat, group = group)) +  # Overlay U.S. state boundaries using map data.
  facet_grid(~ date) +  # Create separate panels for each date.
  coord_fixed(xlim = c(-105, -75), ylim = c(25, 50)) +  # Fix the aspect ratio and limit the map to 
  #the specified longitude and latitude range.
  xlab("Longitude (deg)") + ylab("Latitude (deg)") +  # Label the x- and y-axes.
  theme_bw()  # Apply a black-and-white theme.

NOAA_plot  # Display the completed plot.

```

## Time-series plots — one location, many time points

A **time-series plot** shows how a value changes over time, at **one location** (or several).

Let's pick 8 stations at random and look at their daily maximum temperatures throughout time:

```{r time-series, fig.height=4}
set.seed(12)
UIDs      <- unique(Tmax$id)  # Extract the unique station IDs from the Tmax dataset.
UIDs_sub  <- sample(UIDs, 8)  # Randomly select 8 station IDs.
Tmax_sub  <- filter(Tmax, id %in% UIDs_sub)  # Subset the data to include only 
# the selected station IDs.

TmaxTS <- ggplot(Tmax_sub) +  # Create a ggplot object using the subsetted data.
  geom_line(aes(x = t, y = z)) +  # Add a line plot of temperature (z) over time (t).
  facet_wrap(~ id, ncol = 4) +  # Create separate panels for each station ID arranged in 5 columns.
  xlab("Day number (from May 1, 2023)") +  # Label the x-axis.
  ylab("Tmax (degC)") +  # Label the y-axis.
  theme_bw() +  # Apply a black-and-white theme.
  theme(panel.spacing = unit(1, "lines"))  # Adjust spacing between panels.

TmaxTS  # Display the completed time series plot.
```

## Hovmöller plots — space vs. time in one picture

Sometimes we want to see *both* space and time in **one 2-D plot**. A **Hovmöller plot** does exactly that: one axis is space (say, latitude), the other axis is time, and color is the value.

Because our stations are irregular in space, we first bin them into latitude bands and average before plotting:

```{r hovmoller}
lim_lat  <- range(Tmax$lat)  # Compute the minimum and maximum latitude values.
lim_t    <- range(Tmax$t)  # Compute the minimum and maximum time values.
lat_axis <- seq(lim_lat[1], lim_lat[2], length = 25)  # Create 25 evenly spaced latitude grid points.
t_axis   <- seq(lim_t[1],  lim_t[2],  length = 100)  # Create 100 evenly spaced time grid points.

Tmax_grid       <- Tmax  # Make a copy of the original dataset.
dists           <- abs(outer(Tmax$lat, lat_axis, "-"))  
# Compute distances between each latitude and grid latitude.
Tmax_grid$lat   <- lat_axis[apply(dists, 1, which.min)]  
# Replace each latitude with its nearest grid latitude.

Tmax_lat_Hov <- group_by(Tmax_grid, lat, t) %>%  # Group data by grid latitude and time.
  summarise(z = mean(z, na.rm = TRUE), .groups = "drop")  
# Compute the mean temperature for each latitude–time pair.

Hovmoller_lat <- ggplot(Tmax_lat_Hov) +  # Create a ggplot object using the aggregated data.
  geom_tile(aes(x = lat, y = t, fill = z)) +  # Draw a heatmap with latitude on x, 
  # time on y, and color representing temperature.
  scale_fill_distiller(name = "degC", palette = "Spectral") +  
  # Apply a Spectral color palette labeled "degC".
  xlab("Latitude (deg)") + ylab("Day number (days)") +  # Label the axes.
  theme_bw()  # Apply a black-and-white theme.

Hovmoller_lat  # Display the Hovmöller diagram.
```



> **In-class exercise:** Can you produce a similar plot with latitude replaced by longitude?  What do you expect to see?

```{r hovmoller}



```



# PART 3 — Numerical Summaries for Spatio-Temporal Data

## Empirical Spatial Mean

This gives us **one number per station** — the typical temperature at that station over the summer.

```{r spat-mean}
spat_av <- group_by(Tmax, lat, lon) %>%  # Group the data by 
  # each spatial location (latitude and longitude).
  summarise(mu_emp = mean(z, na.rm = TRUE), .groups = "drop")  
# Compute the mean temperature at each location and drop grouping.

head(spat_av)  # Display the first few rows of the spatial averages dataset.
```

Now plot the spatial mean vs. latitude and vs. longitude:

```{r spat-mean-plot, fig.height=2, fig.width=4}
library(patchwork)  # Load patchwork to combine multiple ggplots.

lat_means <- ggplot(spat_av) +  # Create a scatter plot using the spatial averages data.
  geom_point(aes(lat, mu_emp)) +  # Plot mean temperature against latitude.
  xlab("Latitude (deg)") +  # Label the x-axis.
  ylab("Mean Tmax (degC)") + theme_bw()  # Label the y-axis and apply a clean theme.

lon_means <- ggplot(spat_av) +  # Create another scatter plot using the same data.
  geom_point(aes(lon, mu_emp)) +  # Plot mean temperature against longitude.
  xlab("Longitude (deg)") +  # Label the x-axis.
  ylab("Mean Tmax (degC)") + theme_bw()  # Label the y-axis and apply a clean theme.

lat_means + lon_means  # Display the two plots side by side.
```




> **In-class exercise:** Can we study the minimum temperature similarly?  What do you see from the plots?

```{r}







```



## Empirical Temporal Mean

This gives **one number per day** — the average temperature over the whole region on that day.

```{r time-mean}
Tmax_av <- group_by(Tmax, date) %>%  # Group the data by date.
  summarise(meanTmax = mean(z, na.rm = TRUE), .groups = "drop")  
# Compute the average maximum temperature for each date.

ggplot() +  # Initialize an empty ggplot object.
  geom_line(data = Tmax,
            aes(x = date, y = z, group = id),
            colour = "blue", alpha = 0.04) +  # Plot individual station time series with transparency.
  geom_line(data = Tmax_av,
            aes(x = date, y = meanTmax), linewidth = 0.9, group = 1) +  
  # Overlay the overall daily mean temperature as a thicker line.
  xlab("Date") + ylab("Maximum temperature (degC)") +  # Label the axes.
  theme_bw()  # Apply a clean black-and-white theme.
```

Each blue line is one station; the black line is the daily **regional average**. It clearly rises through spring/early summer, peaks in July/August, and falls in September — the seasonal cycle.

## Spatio-Temporal Semivariogram

### An example

To simplify computations, we use the June 2023 subset.

```{r, warning=FALSE}
library(sp); library(spacetime); library(gstat) 

TmaxJune <- subset(Tmax, month==6) %>%
  select(lat,lon,day,z)
TmaxJune <- na.omit(TmaxJune)

# --- Unique stations and full station × day grid --------------------------
stations <- unique(TmaxJune[, c("lon", "lat")])
rownames(stations) <- NULL
stations$id <- seq_len(nrow(stations))

days <- sort(unique(TmaxJune$day))

full <- expand.grid(id = stations$id, day = days)
full <- merge(full, stations, by = "id")
dat  <- merge(full, TmaxJune, by = c("lon", "lat", "day"), all.x = TRUE)
dat  <- dat[order(dat$day, dat$id), ]     # time-major order (required)
rownames(dat) <- NULL

# --- Spatial points (projected to km) -------------------------------------
sp_pts <- SpatialPoints(stations[, c("lon", "lat")],
                        CRS("+proj=longlat +datum=WGS84"))
sp_pts <- spTransform(sp_pts,
                      CRS("+proj=utm +zone=14 +datum=WGS84 +units=km"))

# --- Time index -----------------------------------------------------------
time_pts <- as.POSIXct(paste0("2023-06-", sprintf("%02d", days)), tz = "UTC")

# --- Build STFDF ----------------------------------------------------------
stfdf <- STFDF(sp   = sp_pts,
               time = time_pts,
               data = data.frame(z = dat$z, lat=dat$lat, lon=dat$lon, day=dat$day))

# --- Empirical semivariogram plot---------------------------------------------
vv <- variogram(z ~ lat + lon + day + I(day^2), data = stfdf,
                width  = 80,          # km
                cutoff = 1000,        # km
                tlags  = 0.01:6.01,   # days
                na.omit = TRUE)

plot(vv, main = "Semivariogram (spatial lag vs temporal lag)")

```


