Below are all the codes for class 1 demostrations


```{r install, eval=FALSE}
install.packages(c("dplyr", "ggplot2", "maps", "patchwork", "tidyverse", "FRK", "lubridate", 
                   "MASS", "gstat", "sp", "spacetime", "animation"))

```

## Our dataset: NOAA daily maximum temperature

```{r}
NOAA_df <- read.csv("NOAA_2023 v2.csv", header=TRUE, sep=",") # read in the dataset
# Create the new "date" column
NOAA_df$date <- as.Date(paste(NOAA_df$year, NOAA_df$month, NOAA_df$day, sep = "-"))
head(NOAA_df)
```

```{r}
library(dplyr) # need to apply functions in the package dplyr and have to load it here
Tmax <- filter(NOAA_df,               # only retrieve the variables and range of time we need
               var  == "Tmax",        # daily maximum temperature
               month %in% 5:9,        # May–September
               year  == 2023)         # year 2023

Tmax$t <- as.numeric(Tmax$date) - as.numeric(as.Date("2023-04-30"))          # day 1 = May 1, 2023

```

---

# PART 1 — Visualizing Spatio-Temporal Data

## Spatial plots — snapshots in time

```{r spatial-plot, fig.height=4}
library(ggplot2) # a useful package for statistical plots
Tmax_1 <- subset(Tmax, t %in% c(1, 32, 62)) # retrieve data on the 3 days only

NOAA_plot <- ggplot(Tmax_1) +  # Create a ggplot object using Tmax_1 as the main dataset.
  geom_point(aes(x = lon, y = lat, colour = z), size = 2) +  # Add points with longitude and latitude, 
  # colored by z, with size 2.
  scale_colour_viridis_c(name = "degC", option = "turbo") +  # Apply a continuous viridis color scale 
  # labeled "degC" using the turbo palette.
  geom_path(data = map_data("state"),
            aes(x = long, y = lat, group = group)) +  # Overlay U.S. state boundaries using map data.
  facet_grid(~ date) +  # Create separate panels for each date.
  coord_fixed(xlim = c(-105, -75), ylim = c(25, 50)) +  # Fix the aspect ratio and limit the map to 
  #the specified longitude and latitude range.
  xlab("Longitude (deg)") + ylab("Latitude (deg)") +  # Label the x- and y-axes.
  theme_bw()  # Apply a black-and-white theme.

NOAA_plot  # Display the completed plot.

```

> **Class discussion:** Where in the map do you see the biggest changes between May 1 and July 1?

## Time-series plots — one location, many time points

A **time-series plot** shows how a value changes over time, at **one location** (or several).

```{r time-series, fig.height=4}
set.seed(12)
UIDs      <- unique(Tmax$id)  # Extract the unique station IDs from the Tmax dataset.
UIDs_sub  <- sample(UIDs, 8)  # Randomly select 8 station IDs.
Tmax_sub  <- filter(Tmax, id %in% UIDs_sub)  # Subset the data to include only 
# the selected station IDs.

TmaxTS <- ggplot(Tmax_sub) +  # Create a ggplot object using the subsetted data.
  geom_line(aes(x = t, y = z)) +  # Add a line plot of temperature (z) over time (t).
  facet_wrap(~ id, ncol = 4) +  # Create separate panels for each station ID arranged in 5 columns.
  xlab("Day number (from May 1, 2023)") +  # Label the x-axis.
  ylab("Tmax (degC)") +  # Label the y-axis.
  theme_bw() +  # Apply a black-and-white theme.
  theme(panel.spacing = unit(1, "lines"))  # Adjust spacing between panels.

TmaxTS  # Display the completed time series plot.
```

## Hovmöller plots — space vs. time in one picture

```{r hovmoller}
lim_lat  <- range(Tmax$lat)  # Compute the minimum and maximum latitude values.
lim_t    <- range(Tmax$t)  # Compute the minimum and maximum time values.
lat_axis <- seq(lim_lat[1], lim_lat[2], length = 25)  # Create 25 evenly spaced latitude grid points.
t_axis   <- seq(lim_t[1],  lim_t[2],  length = 100)  # Create 100 evenly spaced time grid points.

Tmax_grid       <- Tmax  # Make a copy of the original dataset.
dists           <- abs(outer(Tmax$lat, lat_axis, "-"))  
# Compute distances between each latitude and grid latitude.
Tmax_grid$lat   <- lat_axis[apply(dists, 1, which.min)]  
# Replace each latitude with its nearest grid latitude.

Tmax_lat_Hov <- group_by(Tmax_grid, lat, t) %>%  # Group data by grid latitude and time.
  summarise(z = mean(z, na.rm = TRUE), .groups = "drop")  
# Compute the mean temperature for each latitude–time pair.

Hovmoller_lat <- ggplot(Tmax_lat_Hov) +  # Create a ggplot object using the aggregated data.
  geom_tile(aes(x = lat, y = t, fill = z)) +  # Draw a heatmap with latitude on x, 
  # time on y, and color representing temperature.
  scale_fill_distiller(name = "degC", palette = "Spectral") +  
  # Apply a Spectral color palette labeled "degC".
  xlab("Latitude (deg)") + ylab("Day number (days)") +  # Label the axes.
  theme_bw()  # Apply a black-and-white theme.

Hovmoller_lat  # Display the Hovmöller diagram.
```

**How to read it:**

- The **horizontal axis** is latitude — moving right means moving north.
- The **vertical axis** is time — moving *up* means moving forward in time.
- **Color** is temperature.
- We can see the **north (right) is always cooler** than the south, and **temperature peaks in July and August**.

> **In-class exercise:** Can you produce a similar plot with latitude replaced by longitude?  What do you expect to see?  Try it out below:

```{r}

```

## Animations — a movie of your data

The R code below creates an HTML animation of daily maximum temperatures. It may take longer time (a few minutes) and you may run it at home:

```{r animation, fig.width=10, fig.height=7}
library(animation)  # Load the animation package for creating animated outputs.
library(dplyr)  # Load dplyr for data manipulation functions.
library(ggplot2)  # Load ggplot2 for plotting.
library(maps)  # Load maps for geographic boundary data.

# ---- Bounding box: from the data, with generous padding ----
lon_range <- range(Tmax$lon, na.rm = TRUE) + c(-3, 3)  # Compute longitude range and add padding.
lat_range <- range(Tmax$lat, na.rm = TRUE) + c(-2, 2)  # Compute latitude range and add padding.

# ---- Basemaps: countries + US state boundaries ----
world_map  <- map_data("world")  # Retrieve world map polygon data.
states_map <- map_data("state")  # Retrieve US state boundary data.

Tmax_t <- function(tau) {  # Define a function to plot temperature for a given time tau.
  Tmax_sub <- filter(Tmax, t == tau)  # Subset the data for the selected time.

  ggplot() +
    # country polygons (fills Canada / Mexico too so the frame isn't blank)
    geom_polygon(data = world_map,
                 aes(x = long, y = lat, group = group),
                 fill = "grey95", colour = "grey60", size = 0.3) +  # Draw filled country polygons.
    # US state boundaries on top of the country fill
    geom_path(data = states_map,
              aes(x = long, y = lat, group = group),
              colour = "grey55", size = 0.3) +  # Overlay US state boundaries.
    # temperature observations
    geom_point(data = Tmax_sub,
               aes(x = lon, y = lat, colour = z),
               size = 2.5, alpha = 0.9) +  # Plot temperature points colored by z.
    scale_colour_distiller(name = "z",
                           limits = c(-10, 50),
                           palette = "Spectral") +  # Apply a Spectral color scale with fixed limits.
    # Crop to the region and preserve geographic aspect ratio
    coord_quickmap(xlim = lon_range, ylim = lat_range, expand = TRUE) +  
    # Set map extent and aspect ratio.
    labs(title = paste("t =", tau),
         x = "Longitude", y = "Latitude") +  # Add title and axis labels.
    theme_bw(base_size = 13) +  # Apply a clean theme with base font size 13.
    theme(plot.margin = margin(5, 10, 5, 5))  # Adjust plot margins.
}

gen_anim <- function() {  # Define a function to generate animation frames.
  for (t in lim_t[1]:lim_t[2]) {  # Loop over all time values.
    plot(Tmax_t(t))  # Plot each time frame.
  }
}

# ---- Landscape frames sized for the US bounding box ----
ani.options(interval  = 0.2,
            ani.width  = 1000,   # Set animation frame width.
            ani.height = 700)    # Set animation frame height.

saveHTML(gen_anim(),
         autoplay = FALSE, loop = FALSE, verbose = FALSE,
         outdir = ".", htmlfile = "NOAA_anim.html")  # Save the animation as an HTML file.
```

Open `NOAA_anim.html` in your browser to watch the temperature map evolve day by day. You'll often see weather systems **drifting from west to east** — a moving pattern that a static plot cannot easily reveal.

Another example of animation: COVID 19 data animation from the a web page of John's Hopkins University (https://coronavirus.jhu.edu/data/animated-world-map )

---

# PART 2 — Basic Statistical Concepts

## The Mean (Average)

```{r mean-ex}
temps <- c(26, 28, 30, 27, 29)
mean(temps)
```

## The Variance and Standard Deviation


```{r var-ex}
var(temps)        # variance
sd(temps)         # standard deviation
```

```{r var-picture, echo=FALSE, fig.height=3}
df <- data.frame(
  value = c(rnorm(50, 20, 1), rnorm(50, 20, 5)),
  group = rep(c("Small variance", "Large variance"), each = 50)
)
ggplot(df, aes(x = value, fill = group)) +
  geom_histogram(bins = 20, alpha = 0.7) +
  facet_wrap(~ group) +
  theme_bw() + xlab("value") + ylab("count") +
  theme(legend.position = "none")
```

## One-Sample t-Test for the Mean

### Example

Continuing with our 5 temperatures, test whether the true mean differs from 27 °C.

```{r ttest-ex}
temps <- c(26, 28, 30, 27, 29)

# Manual computation
xbar <- mean(temps)
s    <- sd(temps)
n    <- length(temps)
mu0  <- 27

t_stat <- (xbar - mu0) / (s / sqrt(n))
p_val  <- 2 * pt(-abs(t_stat), df = n - 1)   # two-sided

c(t = t_stat, df = n - 1, p_value = p_val)

# Built-in function
t.test(temps, mu = 27)
```

```{r ttest-plot, echo=FALSE, fig.height=3.2, fig.width=6}
library(ggplot2)

# Values from the example above
n       <- 5
df_t    <- n - 1
t_obs   <- (mean(temps) - 27) / (sd(temps) / sqrt(n))
alpha   <- 0.05
t_crit  <- qt(1 - alpha / 2, df = df_t)   # two-sided critical value

# t-density curve
xx   <- seq(-5, 5, length.out = 400)
dens <- data.frame(t = xx, d = dt(xx, df = df_t))

# Two-sided rejection region
rej_left  <- subset(dens, t <= -t_crit)
rej_right <- subset(dens, t >=  t_crit)

# Two-sided p-value region (|T| >= |t_obs|)
pv_left  <- subset(dens, t <= -abs(t_obs))
pv_right <- subset(dens, t >=  abs(t_obs))

ggplot(dens, aes(t, d)) +
  # rejection region (light red)
  geom_area(data = rej_left,  aes(t, d), fill = "red",    alpha = 0.25) +
  geom_area(data = rej_right, aes(t, d), fill = "red",    alpha = 0.25) +
  # p-value region (darker blue)
  geom_area(data = pv_left,   aes(t, d), fill = "steelblue", alpha = 0.55) +
  geom_area(data = pv_right,  aes(t, d), fill = "steelblue", alpha = 0.55) +
  # density curve
  geom_line(size = 0.8) +
  # observed t
  geom_vline(xintercept =  t_obs, colour = "steelblue", linetype = "dashed") +
  geom_vline(xintercept = -t_obs, colour = "steelblue", linetype = "dashed") +
  # critical values
  geom_vline(xintercept =  t_crit, colour = "red", linetype = "dotted") +
  geom_vline(xintercept = -t_crit, colour = "red", linetype = "dotted") +
  annotate("text", x = t_obs,  y = 0.02, label = sprintf("t = %.2f", t_obs),
           colour = "steelblue", hjust = -0.1, size = 3.5) +
  annotate("text", x = t_crit, y = 0.35,
           label = sprintf("t[crit] == %.2f", t_crit),
           parse = TRUE, colour = "red", hjust = -0.1, size = 3.5) +
  labs(
    title    = expression("Student's t-distribution with " * n - 1 == 4 * " df"),
    subtitle = "Red = rejection region (\u03b1 = 0.05, two-sided);  Blue = p-value area",
    x = "t", y = "density"
  ) +
  theme_bw()
```

## Covariance: do two variables move together?

```{r cov-ex}
temp  <- c(26, 28, 30, 27, 29)          # °C
sales <- c(100, 120, 150, 110, 135)     # ice creams
cov(temp, sales)

```

## Correlation: covariance without units

### Correlation gallery

```{r cor-gallery, echo=FALSE, fig.height=3.5}
set.seed(2)
n <- 100
x <- rnorm(n)
sim <- data.frame(
  x = rep(x, 4),
  y = c( x + rnorm(n,0,0.2),           # r ~ +1
        -x + rnorm(n,0,0.2),           # r ~ -1
         x + rnorm(n,0,1.5),           # r ~ +0.5
         rnorm(n))                     # r ~ 0
)
sim$r <- rep(c("r ~ +1", "r ~ -1", "r ~ +0.5", "r ~ 0"), each = n)
ggplot(sim, aes(x, y)) +
  geom_point(alpha = 0.5) + facet_wrap(~ r, nrow = 1) +
  theme_bw()
```

# PART 3 — Numerical Summaries for Spatio-Temporal Data

## Empirical Spatial Mean

This gives us **one number per station** — the typical temperature at that station over the summer.

```{r spat-mean}
spat_av <- group_by(Tmax, lat, lon) %>%  # Group the data by 
  # each spatial location (latitude and longitude).
  summarise(mu_emp = mean(z, na.rm = TRUE), .groups = "drop")  
# Compute the mean temperature at each location and drop grouping.

head(spat_av)  # Display the first few rows of the spatial averages dataset.
```

Now plot the spatial mean vs. latitude and vs. longitude:

```{r spat-mean-plot, fig.height=2, fig.width=4}
library(patchwork)  # Load patchwork to combine multiple ggplots.

lat_means <- ggplot(spat_av) +  # Create a scatter plot using the spatial averages data.
  geom_point(aes(lat, mu_emp)) +  # Plot mean temperature against latitude.
  xlab("Latitude (deg)") +  # Label the x-axis.
  ylab("Mean Tmax (degC)") + theme_bw()  # Label the y-axis and apply a clean theme.

lon_means <- ggplot(spat_av) +  # Create another scatter plot using the same data.
  geom_point(aes(lon, mu_emp)) +  # Plot mean temperature against longitude.
  xlab("Longitude (deg)") +  # Label the x-axis.
  ylab("Mean Tmax (degC)") + theme_bw()  # Label the y-axis and apply a clean theme.

lat_means + lon_means  # Display the two plots side by side.
```

**Observation:** the mean temperature **decreases clearly** with latitude (moving north = cooler). The decreasing trend with longitude is not so strong. So *latitude* is relatively a more useful "explanatory variable" for our data.

> **In-class exercise:** Can we study the minimum temperature similarly?  What do you see from the plots?  Try it below:

```{r}

```


## Empirical Temporal Mean

```{r time-mean}
Tmax_av <- group_by(Tmax, date) %>%  # Group the data by date.
  summarise(meanTmax = mean(z, na.rm = TRUE), .groups = "drop")  
# Compute the average maximum temperature for each date.

ggplot() +  # Initialize an empty ggplot object.
  geom_line(data = Tmax,
            aes(x = date, y = z, group = id),
            colour = "blue", alpha = 0.04) +  # Plot individual station time series with transparency.
  geom_line(data = Tmax_av,
            aes(x = date, y = meanTmax), linewidth = 0.9, group = 1) +  
  # Overlay the overall daily mean temperature as a thicker line.
  xlab("Date") + ylab("Maximum temperature (degC)") +  # Label the axes.
  theme_bw()  # Apply a clean black-and-white theme.
```

## Spatio-Temporal Covariogram and Semivariogram

### An example

To simplify computations, we use the June 2023 subset.

```{r, warning=FALSE}
library(sp); library(spacetime); library(gstat) 

TmaxJune <- subset(Tmax, month==6) %>%
  select(lat,lon,day,z)
TmaxJune <- na.omit(TmaxJune)

# --- Unique stations and full station × day grid --------------------------
stations <- unique(TmaxJune[, c("lon", "lat")])
rownames(stations) <- NULL
stations$id <- seq_len(nrow(stations))

days <- sort(unique(TmaxJune$day))

full <- expand.grid(id = stations$id, day = days)
full <- merge(full, stations, by = "id")
dat  <- merge(full, TmaxJune, by = c("lon", "lat", "day"), all.x = TRUE)
dat  <- dat[order(dat$day, dat$id), ]     # time-major order (required)
rownames(dat) <- NULL

# --- Spatial points (projected to km) -------------------------------------
sp_pts <- SpatialPoints(stations[, c("lon", "lat")],
                        CRS("+proj=longlat +datum=WGS84"))
sp_pts <- spTransform(sp_pts,
                      CRS("+proj=utm +zone=14 +datum=WGS84 +units=km"))

# --- Time index -----------------------------------------------------------
time_pts <- as.POSIXct(paste0("2023-06-", sprintf("%02d", days)), tz = "UTC")

# --- Build STFDF ----------------------------------------------------------
stfdf <- STFDF(sp   = sp_pts,
               time = time_pts,
               data = data.frame(z = dat$z, lat=dat$lat, lon=dat$lon, day=dat$day))

# --- Empirical semivariogram plot---------------------------------------------
vv <- variogram(z ~ lat + lon + day + I(day^2), data = stfdf,
                width  = 80,          # km
                cutoff = 1000,        # km
                tlags  = 0.01:6.01,   # days
                na.omit = TRUE)

plot(vv, main = "Semivariogram (spatial lag vs temporal lag)")

```




