---
title: "CUSA1056 Space-Time Data Science"
author: "Dr. Remus Ho"
date: "`r Sys.Date()`"
subtitle: "Class 3: Regression for spatio-temporal data"
output:
  pdf_document:
    latex_engine: xelatex
    toc: true
    toc_depth: 2
    number_sections: true
geometry: margin=1in
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE,
  fig.align = "center",
  fig.width = 6,      # safe PDF width
  fig.height = 4,
  out.width = "100%", # scale to page width
  fig.crop = FALSE
)
```

# Introduction

In this class we will study **spatio-temporal regression**, i.e. statistical
models for data that vary jointly across space *and* time. We shall consider "trend surface models" only because of its simplicity and as a direct extension from linear regression.

Our roadmap for the morning session today is:

| Part | Topic                                            | 
|:----:|:-------------------------------------------------|
| 1    | Interpolation vs Linear Regression                      |
| 2    | Deterministic Prediction: Inverse Distance Weighting | 
| 3    | Linear Regression Models with Basis Functions           | 
| 4    | Poisson and Negative Binomial Regression for Count Data                | 
---

# Interpolation vs Linear Regression (Part 1)

Before we get to spatial and spatio-temporal models, it is worth spending time on two ideas that underpin almost everything that follows:

- **Interpolation** — passing a smooth (in some sense) function through a set of
  observed points.
- **Linear regression** — the most widely-used framework for turning that
  idea into a *statistical* model, with parameters that we can estimate,
  test, and interpret.

We treat them in parallel because they answer subtly different questions,
and because the distinction will re-appear in Part 2 and 3 when we contrast
**IDW (a pure interpolator)** with **regression-based approaches** to
spatial prediction.

## Interpolation

Suppose we observe pairs \((x_i, y_i)\) for \(i = 1, \ldots, n\) and want a
function \(\widehat{f}(x)\) that describes how \(y\) depends on \(x\). Two
broad strategies exist:

- **Interpolation.** Require the fitted curve to pass *exactly* through
  every observed point:
  $$
  \widehat{f}(x_i) \;=\; y_i \qquad \text{for all } i.
  $$
  Examples: piecewise-linear interpolation, cubic splines, and — in the
  spatial setting — **Inverse Distance Weighting (IDW)**.
- **Regression** Allow the curve to *miss* the observed points in exchange
  for being simpler, more stable, and less sensitive to noise:
  $$
  \widehat{f}(x_i) \;\approx\; y_i,
  $$
  with the discrepancy \(y_i - \widehat{f}(x_i)\) treated as *residual
  variation*. Example: ordinary least-squares regression.

The right choice depends on whether we believe the data are essentially
**noise-free** (interpolation) or **noisy** (regression). For most
real-world problems — certainly for COVID-19 case counts, temperature
observations, or house prices — the data contain substantial noise, and a
more statistical approach is more honest.

We now focus on the very important **linear
regression** method.

## Linear Regression

### The Model

Given observations \((\mathbf{x}_i, y_i)\) for \(i = 1, \ldots, n\), where
\(\mathbf{x}_i = (x_{i1}, \ldots, x_{ip})^{\top}\), the linear regression
model assumes

$$
y_i \;=\; \beta_0 + \beta_1 x_{i1} + \cdots + \beta_p x_{ip} + \varepsilon_i,
\qquad \varepsilon_i \stackrel{\text{iid}}{\sim} N(0, \sigma^{2}).
$$

The word "linear" refers to *linearity in the coefficients* \(\beta_k\),
**not** in the covariates. This is a crucial and often-missed point: we can
use non-linear transformations of \(x\) (squares, logs, sines, radial basis
functions) as columns of our design matrix, and the model is still linear.
This observation is the seed of the basis-function expansion we will
develop in Part 3.

### Matrix Form and the Design Matrix

Stacking all \(n\) equations gives the compact form

$$
\mathbf{y} \;=\; \mathbf{X}\boldsymbol{\beta} + \boldsymbol{\varepsilon},
$$

where

$$
\mathbf{y} = \begin{pmatrix} y_1 \\ \vdots \\ y_n \end{pmatrix},
\quad
\mathbf{X} = \begin{pmatrix}
1 & x_{11} & \cdots & x_{1p} \\
\vdots & \vdots & \ddots & \vdots \\
1 & x_{n1} & \cdots & x_{np}
\end{pmatrix},
\quad
\boldsymbol{\beta} = \begin{pmatrix} \beta_0 \\ \vdots \\ \beta_p \end{pmatrix}.
$$

The matrix \(\mathbf{X}\) is the **design matrix**, and its columns are the
**features** we hand to the regression. When we replace the raw \(x_{ij}\)
with basis functions \(\phi_k(\mathbf{s}_i)\) in Part 3, we will simply be
building a different design matrix.

### Ordinary Least Squares (OLS)

OLS chooses \(\boldsymbol{\beta}\) to minimise the sum of squared residuals,

$$
\widehat{\boldsymbol{\beta}}
\;=\; \arg\min_{\boldsymbol{\beta}}
\sum_{i=1}^{n} \bigl(y_i - \mathbf{x}_i^{\top}\boldsymbol{\beta}\bigr)^{2}
\;=\; \arg\min_{\boldsymbol{\beta}}
\|\mathbf{y} - \mathbf{X}\boldsymbol{\beta}\|^{2}.
$$

Setting the gradient to zero yields the celebrated **normal equations**,
whose solution is the closed form

$$
\widehat{\boldsymbol{\beta}} \;=\; (\mathbf{X}^{\top}\mathbf{X})^{-1}\mathbf{X}^{\top}\mathbf{y}.
$$

Fitted values and residuals are

$$
\widehat{\mathbf{y}} = \mathbf{X}\widehat{\boldsymbol{\beta}}, \qquad
\mathbf{e} = \mathbf{y} - \widehat{\mathbf{y}}.
$$

### Interpreting the Coefficients

Each coefficient \(\beta_k\) has a "partial-effect" interpretation:

> \(\beta_k\) is the expected change in \(y\) associated with a **one-unit
> increase in \(x_k\)**, holding all other covariates fixed.

The intercept \(\beta_0\) is the expected value of \(y\) when all covariates
are zero. Statistical software also reports:

- **Standard errors** \(\mathrm{SE}(\widehat{\beta}_k)\), quantifying
  sampling uncertainty;
- **t-statistics** \(t_k = \widehat{\beta}_k / \mathrm{SE}(\widehat{\beta}_k)\)
  and their **p-values**, testing \(H_0: \beta_k = 0\);
- **\(R^{2}\)**, the fraction of variance in \(y\) explained by the model.

### Key Assumptions

The OLS coefficient estimates are unbiased and have the standard sampling
distribution provided:

- **Linearity**;
- **Independent** errors \(\varepsilon_i\);
- **Constant variance** (homoscedasticity), \(\mathrm{Var}[\varepsilon_i] = \sigma^{2}\);
- **Normal** errors.

While linear regression is very useful, the above assumptions may fail in some applications.  For example in part 4, **Poisson / Negative Binomial GLMs** will be used to handle non-Gaussian responses.

## Numerical Example: A Simple Simulated Dataset

To make the mechanics concrete, we simulate 40 points from the true
relationship

$$
y_i \;=\; 2 + 0.8\, x_i + \varepsilon_i,
\qquad \varepsilon_i \sim N(0, 1.2^{2}),
$$

fit an OLS line, and inspect the output.

```{r ols-example, fig.width=7, fig.height=4}
set.seed(2026)        # Set the random seed to 2026 to make the random number generation reproducible.

n <- 40               # Define the sample size n as 40 observations.

x <- runif(n, min = 0, max = 10)  
# Generate n random values for x from a uniform distribution between 0 and 10.

y <- 2 + 0.8 * x + rnorm(n, sd = 1.2)  
# Generate y values using a linear model with intercept 2 and slope 0.8 plus normally distributed 
# noise with standard deviation 1.2.

# Fit OLS
fit <- lm(y ~ x)      # Fit an ordinary least squares regression model of y on x.

summary(fit)          # Display a detailed summary of the fitted regression 
                      # model, including coefficients and statistics.
```

Inspecting the output, the estimated coefficients should be close to the
true values \(\beta_0 = 2\) and \(\beta_1 = 0.8\), with standard errors
that reflect the sample size and the noise level \(\sigma\).

### Visualising the Fit

```{r ols-plot, fig.width=7, fig.height=4}
plot(x, y, pch = 16, col = "steelblue",
     xlab = "x", ylab = "y",
     main = "OLS line with residuals")  
# Create a scatter plot of x and y with solid blue points, labeled axes, and a descriptive title.

abline(fit, col = "darkred", lwd = 2)  
# Add the fitted OLS regression line to the plot in dark red with a thicker line width.

# Draw residuals as vertical segments
y_hat <- fitted(fit)  
# Compute the fitted (predicted) values from the regression model and store them in y_hat.

segments(x, y, x, y_hat, col = "grey60", lty = 2)  
# Draw dashed grey vertical lines from each observed y value to its corresponding fitted value to
# #visualize residuals.

legend("topleft", bty = "n", cex = 0.9,
       legend = c("data", "OLS fit", "residuals"),
       col    = c("steelblue", "darkred", "grey60"),
       pch    = c(16, NA, NA),
       lty    = c(NA, 1, 2),
       lwd    = c(NA, 2, 1))  
# Add a legend in the top-left corner explaining the data points, OLS line, and residual lines with
# matching styles.
```

Notice two things in the plot:

- The OLS line **does not pass through every point** — this is not interpolation. 
- The residuals (grey dashed segments) are what OLS minimises: the sum of
  their squared lengths.

### Reconstructing OLS "by Hand"

To reinforce the matrix formulation, we can compute
\(\widehat{\boldsymbol{\beta}} = (\mathbf{X}^{\top}\mathbf{X})^{-1}\mathbf{X}^{\top}\mathbf{y}\)
directly and confirm we recover the same numbers as `lm()`.

```{r ols-by-hand}
X <- cbind(1, x)                              # design matrix
beta_hat <- solve(t(X) %*% X) %*% t(X) %*% y  # normal equations
beta_hat
coef(fit)                                     # compare with lm()
```

The two sets of numbers agree to machine precision. 

### A Non-Linear function Extension: Same Machinery, New Columns

To make the "linear in the coefficients" point tangible, we now fit a
**quadratic** curve to a slightly non-linear dataset, using OLS with a
design matrix whose columns are \((1, x, x^{2})\).

```{r ols-quadratic, fig.width=7, fig.height=4}
set.seed(7)  
# Reproducibility

x  <- seq(0, 10, length.out = 60)  
# x grid (0–10, 60 pts)

y  <- 1 + 0.5 * x - 0.05 * x^2 + rnorm(length(x), sd = 0.6)  
# Quadratic model + noise

fit_quad <- lm(y ~ x + I(x^2))  
# Fit quadratic regression

coef(fit_quad)  
# Show coefficients

# Plot the fit
xg <- seq(0, 10, length.out = 200)  
# Fine grid for smooth curve

yg <- predict(fit_quad, newdata = data.frame(x = xg))  
# Predicted values

plot(x, y, pch = 16, col = "steelblue",
     xlab = "x", ylab = "y",
     main = "Quadratic fit via linear regression")  
# Scatter plot

lines(xg, yg, col = "darkred", lwd = 2)  
# Add fitted curve
```

The **fitted curve is non-linear**, yet the *model* is a linear regression
— we merely added a column \(x^{2}\) to the design matrix. This is exactly
the trick that we will exploit in Part 3 to build flexible spatial
regressions on top of simple linear machinery.

With this foundation in place, we can turn in Part 2 to our first
spatial method — **Inverse Distance Weighting** — and see how the
interpolation method being applied on a map.
---

# Deterministic Prediction — Inverse Distance Weighting (Part 2)

## Motivation

Consider the NOAA data set of **maximum daily temperatures** in July 2023 that we have studied in class 1. Suppose we want to predict the
temperature at a location \(\mathbf{s}_0\) on 15 July 2023 where no observation (or even no station) exists.

*How should we combine the observations?* A natural intuition is
**Tobler's First Law of Geography**:

> *"Everything is related to everything else, but near things are more related
> than distant things."*

This suggests giving **more weight** to observations that are *close* to the
prediction location — and because we have data at multiple times, "close"
should be understood in both **space** and **time**.

## Inverse Distance Weighting (IDW)

Suppose the data are
$$
\{Z(\mathbf{s}_{ij}; t_j) : i = 1, \ldots, m_j;\ j = 1, \ldots, T\}.
$$

The **inverse-distance-weighted (IDW) predictor** at a location
\((\mathbf{s}_0; t_0)\) is a weighted average of *all* observations:

$$
\widehat{Z}(\mathbf{s}_0; t_0) \;=\; \sum_{j=1}^{T} \sum_{i=1}^{m_j}
   w_{ij}(\mathbf{s}_0; t_0)\, Z(\mathbf{s}_{ij}; t_j),
$$

where the weights are

$$
w_{ij}(\mathbf{s}_0; t_0) \;=\;
\frac{\widetilde{w}_{ij}(\mathbf{s}_0; t_0)}
     {\sum_{k=1}^{T}\sum_{\ell=1}^{m_k}\widetilde{w}_{\ell k}(\mathbf{s}_0; t_0)},
\qquad
\widetilde{w}_{ij}(\mathbf{s}_0; t_0) \;=\;
\frac{1}{d\bigl((\mathbf{s}_{ij}; t_j),(\mathbf{s}_0; t_0)\bigr)^{\alpha}}.
$$

Key features:

- \(d(\cdot,\cdot)\) is a distance function (often Euclidean over
  \((\text{lon},\text{lat},t)\)).
- \(\alpha > 0\) is the **power parameter**; larger \(\alpha\) concentrates the
  weight on the nearest observations (less smoothing), smaller \(\alpha\)
  gives a smoother surface.
- The weights sum to 1, so \(\widehat{Z}\) is a **weighted average**.
- If \((\mathbf{s}_0; t_0)\) coincides with a data location, then
  \(\widetilde{w}\to\infty\) there and the predictor **interpolates exactly**.

## IDW in R — using `gstat`

We will interpolate temperatures on a spatio-temporal grid, deliberately
**omitting 15 July 2023** from the training data, so that we can later see
how the method behaves on a day with no observations.

```{r idw-gstat}
library("dplyr")
library("ggplot2")
library("gstat")

NOAA_df <- read.csv("NOAA_2023 v2.csv", header = TRUE, sep = ",")  
# Load NOAA data from CSV

# Subset to Tmax, July 2023
Tmax <- filter(NOAA_df,
               var  == "Tmax" &
               month == 7     &
               year  == 2023)  
# Filter: Tmax, July 2023

# 20 x 20 spatial grid, 6 time points across the month
pred_grid <- expand.grid(lon = seq(-100, -80, length = 20),
                         lat = seq(  32,  46, length = 20),
                         day = seq(   5,  30, length = 6))  
# Create space–time prediction grid

# Remove 15 July from training data
Tmax_no_15 <- filter(Tmax, !(day == 15))  
# Exclude day 15 observations

# IDW with alpha = 4
Tmax_July_idw <- idw(formula   = z ~ 1,
                     locations = ~ lon + lat + day,
                     data      = Tmax_no_15,
                     newdata   = pred_grid,
                     idp       = 4)  
# Perform 3D IDW interpolation (power = 4)
```

The main arguments of `gstat::idw` are:

- `formula`   — the variable to interpolate (here `z ~ 1`, i.e. no covariates).
- `locations` — the spatial and temporal coordinates.
- `data`      — the observed data.
- `newdata`   — the prediction locations (space–time grid).
- `idp`       — the power parameter \(\alpha\).

We can then plot `Tmax_July_idw$var1.pred` on the prediction grid.
Note that on 15 July the predicted surface will look **smoother** than on
the surrounding days: with no data available on that day, the predictor
borrows information from adjacent days. This is a typical behaviour of any
predictor on days with no observations.

```{r}
library("dplyr")      # Data manipulation
library("ggplot2")    # Plotting
library("gstat")      # Spatial interpolation (IDW)
library("maps")       # Map outlines

# ---------------------------------------------------------------------------
# 1. Attach predictions to the grid
# ---------------------------------------------------------------------------

pred_grid1 <- pred_grid %>%
  mutate(z_pred = Tmax_July_idw$var1.pred)  
# Add IDW predictions to grid

pred_days <- sort(unique(pred_grid1$day))  
# Get the 6 prediction days

Tmax_days <- Tmax %>% 
  filter(day %in% pred_days)  
# Keep observations for those days

# ---------------------------------------------------------------------------
# 2. Basemap (state + country outlines)
# ---------------------------------------------------------------------------

states_map <- map_data("state")  
# Load US state boundaries

world_map  <- map_data("world")  
# Load world country boundaries

lon_range <- range(pred_grid1$lon)  
# Longitude limits

lat_range <- range(pred_grid1$lat)  
# Latitude limits

# ---------------------------------------------------------------------------
# 3. Faceted map: one panel per day
# ---------------------------------------------------------------------------

ggplot() +

  geom_raster(data = pred_grid1,
              aes(x = lon, y = lat, fill = z_pred)) +
  # IDW surface (filled grid)

  geom_polygon(data = world_map,
               aes(x = long, y = lat, group = group),
               fill = NA, colour = "black", size = 0.3) +
  # Country borders

  geom_path(data = states_map,
            aes(x = long, y = lat, group = group),
            colour = "grey40", size = 0.25) +
  # State borders

  geom_point(data = Tmax_days,
             aes(x = lon, y = lat, fill = z),
             shape = 21, colour = "white", size = 1.8, stroke = 0.3) +
  # Observed stations

  scale_fill_distiller(name = expression(T[max]~"("*degree*C*")"),
                       palette = "Spectral",
                       limits  = c(10, 45)) +
  # Color scale for temperature

  coord_quickmap(xlim = lon_range, ylim = lat_range, expand = FALSE) +
  # Fix map extent and aspect ratio

  facet_wrap(~ day, nrow = 2,
             labeller = labeller(day = function(d) paste("July", d, "2023"))) +
  # One panel per day

  labs(x = "Longitude", y = "Latitude",
       title = expression("IDW interpolation of "*T[max]*
                          " (July 2023), "*alpha*" = 4"),
       subtitle = "15 July was withheld from the training data") +
  # Titles and labels

  theme_bw(base_size = 12) +
  # Clean theme

  theme(strip.background = element_rect(fill = "grey90"),
        panel.grid.minor = element_blank())  
  # Minor styling adjustments
```




## Choosing \(\alpha\): Cross-Validation

The parameter \(\alpha\) controls the **degree of smoothing**. Its value
should not be picked by eye — the standard objective tool is
**cross-validation (CV)**.

### K-fold Cross-Validation

Randomly split the data into \(K\) roughly equal folds. For each fold
\(k = 1,\ldots,K\):

1. Fit the model on the remaining \(K-1\) folds.
2. Predict the observations in fold \(k\).
3. Compute the mean squared prediction error
   $$
   \mathrm{MSPE}_k \;=\; \frac{1}{m_k}\sum_{i=1}^{m_k}\bigl(Z_i - \widehat Z_i^{(-k)}\bigr)^2.
   $$

The K-fold CV score is

$$
CV_{(K)} \;=\; \frac{1}{K}\sum_{k=1}^{K}\mathrm{MSPE}_k .
$$

Common choices are \(K=5\) or \(K=10\).

### Leave-One-Out CV (LOOCV)

The special case \(K = m\) is **leave-one-out CV**:

$$
CV_{(m)} \;=\; \frac{1}{m}\sum_{i=1}^{m}\bigl(Z_i - \widehat Z_i^{(-i)}\bigr)^2.
$$

LOOCV has **low bias** but can have **high variance**, and it can be
computationally expensive since it requires \(m\) refits.

### Tuning \(\alpha\) by LOOCV in R

The idea is simple: for each candidate value of \(\alpha\), do IDW using
each observation in turn as a held-out point, compute the squared error, and
average. Then pick the \(\alpha\) that minimises the CV score.

We can do this by calling `gstat::idw` in a loop and using the observation
locations themselves as `newdata` (leaving one out at a time).

```{r loocv-idw}
# Candidate values of alpha
alphas <- seq(2, 8, by = 1)
CV     <- numeric(length(alphas))

n <- nrow(Tmax)

for (a in seq_along(alphas)) {
  sq_err <- numeric(n)

  for (i in 1:n) {
    # leave observation i out
    train <- Tmax[-i, ]
    test  <- Tmax[ i, ]

    pred <- idw(formula   = z ~ 1,
                locations = ~ lon + lat + day,
                data      = train,
                newdata   = test,
                idp       = alphas[a],
                debug.level = 0)

    sq_err[i] <- (test$z - pred$var1.pred)^2
  }

  CV[a] <- mean(sq_err)
}

# LOOCV curve and optimum
plot(alphas, CV, type = "o",
     xlab = expression(alpha),
     ylab = expression(CV[(m)](alpha)))

alphas[which.min(CV)]   # optimal alpha
min(CV)                 # minimum CV score
```

**Interpretation.** For the NOAA July 2023 data, the LOOCV curve is
convex with a clear minimum around \(\alpha \approx 4\), giving a minimum
CV score of about 4.23. This tells us that giving fairly strong weight to
nearby observations works best for these data.

> **Practical tip.** For a large dataset, LOOCV can be slow because it
> refits \(m\) times. A common compromise is **10-fold CV**, which is much
> faster and empirically gives a good bias–variance trade-off.

Limitations of Deterministic Predictors: IDW is simple and fast, but it does not explicitly model **measurement error** and provides **no model-based prediction variance**. These limitations motivate the **statistical models** we look at next.

---

# Linear Regression Models with Basis Functions (Part 3)

## From Deterministic to Statistical Prediction

Instead of averaging observations, we can specify a **regression model** that
attempts to capture spatio-temporal structure through **covariates** (a
"trend surface"). Formally, for observation \(Z(\mathbf{s}_i; t_j)\):

$$
Z(\mathbf{s}_i; t_j) \;=\; \beta_0 + \beta_1 X_1(\mathbf{s}_i; t_j)
                            + \cdots + \beta_p X_p(\mathbf{s}_i; t_j)
                            + e(\mathbf{s}_i; t_j),
$$

with \(e(\mathbf{s}_i; t_j) \stackrel{\text{iid}}{\sim} N(0,\sigma_e^2)\).

Advantages over deterministic predictors:

- Easy to implement in any regression software.
- Provides **model-based prediction standard errors**.
- Allows **inference** on covariate effects.

## Basis Functions

### Motivation

A plain linear regression in the coordinates,
$$
Y(\mathbf{s}) \;=\; \beta_0 + \beta_1 s_1 + \beta_2 s_2,
$$
can only produce a **tilted plane** as a fitted surface. Real
spatio-temporal processes (temperature fields, infection rates, pollution
levels) are rarely that simple — they exhibit bumps, ridges, and localised
features.

The idea of a **basis-function expansion** is to enrich the design matrix
with a set of *known* functions of the coordinates, so that a *linear* model
in the coefficients can represent a highly *non-linear* surface. Formally,
we write

$$
Y(\mathbf{s}) \;=\; \alpha_1 \phi_1(\mathbf{s}) + \alpha_2 \phi_2(\mathbf{s})
                 + \cdots + \alpha_r \phi_r(\mathbf{s})
              \;=\; \sum_{k=1}^{r} \alpha_k \phi_k(\mathbf{s}),
$$

where:

- \(\phi_1,\ldots,\phi_r\) are **known** basis functions chosen in advance;
- \(\alpha_1,\ldots,\alpha_r\) are **unknown coefficients** to be estimated
  (by OLS, once the \(\phi_k(\mathbf{s}_i)\) are treated as columns of a
  design matrix).

### Two Families: Global vs Local

Basis functions fall broadly into two families:

- **Global basis functions** are non-zero over the *entire* domain. Changing
  the coefficient of one such basis therefore affects the fitted surface
  *everywhere*. Examples: **polynomials** and **sines/cosines** (Fourier basis).
- **Local basis functions** are non-zero only in a small neighbourhood of a
  *centre* (or are so strongly concentrated that they may as well be).
  Changing the coefficient of one such basis affects the fit only *near*
  its centre. Examples: **bisquare** and **wavelets**, and — in practice — **Gaussian radial basis functions**.

We now illustrate these two ideas with the two examples in the title.

### Example 1: Polynomial Basis (Global)

The polynomial basis of order \(p\) on a one-dimensional domain is

$$
\phi_0(s) = 1,\quad \phi_1(s) = s,\quad \phi_2(s) = s^2,\quad \ldots,\quad
\phi_p(s) = s^p.
$$

Each \(\phi_k(s) = s^k\) is non-zero for every \(s \ne 0\), and in fact grows
without bound as \(|s|\) becomes large. This is the defining property of a
**global** basis: the function has support over the whole real line.

Consequences:

- Fitting a polynomial regression is a *global* operation — a single
  observation in one part of the domain can influence the fit in another
  part.
- Increasing the order \(p\) gives more flexibility, but high-order
  polynomials are notoriously unstable.

### Example 2: Gaussian Radial Basis (Local, in practice)

A **Gaussian radial basis function** centred at \(\mathbf{c}_k\) with
bandwidth \(\theta\) is

$$
\phi_k(\mathbf{s}) \;=\; \exp\!\Bigl(-\tfrac{\|\mathbf{s} - \mathbf{c}_k\|^2}{\theta^2}\Bigr).
$$

To build a basis expansion we place a *grid* of centres
\(\mathbf{c}_1, \ldots, \mathbf{c}_r\) over the domain and use one Gaussian
"bump" per centre.

Strictly speaking, a Gaussian is *never exactly zero*, so it is not
technically compactly supported. However, it decays so rapidly that its
value is negligible more than a few bandwidths away from its centre. For
this reason, Gaussian radial basis functions behave, in practice, as **local**
basis functions: changing \(\alpha_k\) mostly affects the fit near
\(\mathbf{c}_k\). Truly local bases such as **bisquare** are exactly zero
outside a fixed radius.

### Visualising the Two Bases (1D)

The following R snippet plots a degree-4 polynomial basis and a set of
Gaussian radial basis functions on the interval \([0, 1]\), to make the
global vs local distinction visible.

```{r basis-illustration, fig.width=8, fig.height=4}
op <- par(mfrow = c(1, 2))

s <- seq(0, 1, length = 400)

# --- Polynomial basis (global) ---
matplot(s, cbind(1, s, s^2, s^3, s^4), type = "l", lty = 1, lwd = 2,
        xlab = "s", ylab = expression(phi[k](s)),
        main = "Polynomial basis (global)")
legend("topleft",
       legend = c("1", "s", "s^2", "s^3", "s^4"),
       col = 1:5, lty = 1, bty = "n", cex = 0.8)

# --- Gaussian radial basis (local in practice) ---
centres <- seq(0.1, 0.9, by = 0.1)
theta   <- 0.08
Phi     <- sapply(centres, function(c) exp(-((s - c)^2) / theta^2))

matplot(s, Phi, type = "l", lty = 1, lwd = 2,
        xlab = "s", ylab = expression(phi[k](s)),
        main = "Gaussian radial basis (local)")

par(op)
```

Looking at the two panels:

- Each **polynomial** curve spans the *whole* interval and its magnitude is
  meaningful at every \(s\). A single coefficient therefore has global
  influence.
- Each **Gaussian bump** is concentrated around its centre and effectively
  zero elsewhere. A single coefficient influences the fit only in a small
  neighbourhood of that centre.

### From 1D to 2D (Space) and to Space–Time

The same ideas extend directly to two dimensions. A Gaussian radial basis
on the plane,
$$
\phi_k(\mathbf{s}) \;=\; \exp\!\Bigl(-\tfrac{\|\mathbf{s} - \mathbf{c}_k\|^2}{\theta^2}\Bigr),
\qquad \mathbf{s} \in \mathbb{R}^2,
$$
gives a smooth "hill" centred at \(\mathbf{c}_k\) — exactly what we need to
represent localised spatial features. For space–time problems we can form
**tensor products** (too technical to explain here, understand it conceptually is sufficient) of spatial and temporal bases,
\(\phi_{k}(\mathbf{s})\,\psi_{\ell}(t)\), to describe features that are
localised in both space and time.

### Choosing a Basis in Practice

Two practical points:

- **Number and placement of centres.** For radial bases we need to choose
  how many bumps to use and where to place them. A common strategy is a
  regular grid over the domain, possibly at *multiple resolutions*
  (coarse + fine), so that the fit can pick up features of different sizes.
  The `FRK` package (used later) automates this with `auto_basis()`.
- **Bandwidth \(\theta\).** Small \(\theta\) gives narrow, very local bumps
  (wiggly fit); large \(\theta\) gives wide bumps (smooth fit). Just as with
  \(\alpha\) in IDW, this is a smoothing parameter that can be tuned by
  cross-validation.

With this vocabulary in place, we now turn to application to a new COVID-19
example, where a linear trend surface is combined with radial basis
functions to capture residual spatial structure.

## Application to a dataset of COVID cases in US

The dataset used for the regression combines publicly available sources at the US county level. Daily cumulative COVID-19 case and death counts were obtained from the Johns Hopkins University Center for Systems Science and Engineering (JHU CSSE) county-level time series, from which weekly new cases and new deaths were derived. The series covers every week from 22 January 2020, when JHU began reporting, through 9 March 2023, when the county-level feed was retired, giving just over three years of weekly observations that span all major waves of the US epidemic. County boundaries and demographic information were obtained from the 2016–2020 five-year American Community Survey (ACS) via the US Census Bureau's API, specifically table S1810, which provides the percentage of the civilian non-institutionalised population reporting a disability — used here as a proxy for the share of long-term ill residents. Longitude and latitude were computed as the geographic centroids of each county's polygon. 

The sources were merged on the five-digit county FIPS code, yielding an analytic dataset in which each row corresponds to one county on one week, with variables for new cases, new deaths, centroid longitude and latitude, weeks elapsed since the start of the series, and the county's share of population with a disability.

Fitting is by ordinary least squares (OLS) that picks parameters that minimize RSS:

$$
\mathrm{RSS} \;=\; \sum_{j=1}^{T}\sum_{i=1}^{m}
   \bigl(Z(\mathbf{s}_i; t_j) - \widehat Z(\mathbf{s}_i; t_j)\bigr)^2,
$$

with \(\widehat\sigma_e^2 = \mathrm{RSS}/(mT - p - 1)\).

Once fitted we can **predict** \(Z(\mathbf{s}_0; t_0)\) at any location for
which we have covariates, together with a **prediction standard error**.
Because our basis functions here are spatio-temporally smooth, the fitted
surface is much smoother than the IDW predictions.

## Constructing Basis Functions with `FRK`

The **FRK** package can automatically build basis functions at multiple
resolutions.

```{r frk-basis}
library("FRK")
library("sp")

covid19_US <- read.csv("covid_US_weekly.csv", header=T, sep=",")
covid19_US$new_cases_per_ten_thousand <-covid19_US$new_cases/covid19_US$population*100000

# `covid19_US` is our spatio-temporal data frame with columns

# Build Gaussian radial basis functions at a single resolution
RB <- auto_basis(data = covid19_US[, c("lon", "lat")] %>% SpatialPoints(),
                nres = 1,
                type = "Gaussian")

# Evaluate basis functions at each observation location
RB_col <- eval_basis(basis = RB,
                s = covid19_US[, c("lon", "lat")] %>% as.matrix()) %>% as.matrix()

colnames(RB_col) <- paste0("B", 1:ncol(RB_col))
```

The columns of `RB_col` are simply new covariates (one per basis function) that
we can plug into a linear model.

## Application: COVID-19 Regression Across US counties

Our response variable is the weekly number of new cases per 100000 people in a county.  The independent variables are latitude, longitude, week number, basis function and one more covariate, the percentage of longterm illness in the county.

### Attach Basis Functions

```{r covid-basis}
covid19_reg <- cbind(covid19_US, RB_col) %>%
  dplyr::select(new_cases_per_ten_thousand,
                lon, lat, week, pct_disab,
                B1:B9)
```

### Fit the regression model

```{r covid-lm}
library(lubridate)

m1_eq  <- new_cases_per_ten_thousand ~ lon + lat + week + pct_disab + B1 + 
            B2 + B3 + B4 + B5 + B6 +B7 +B8 +B9

m1 <- lm(formula = m1_eq,
           data    = covid19_reg)

summary(m1)
```

### Interpretation

- **`pct_disab`** — a one-unit increase in the share of long-term ill
  population is associated with a large increase in the rate of new COVID-19
  cases per 100,000 people. The coefficient is highly significant, in line
  with the exploratory findings that vulnerable populations experienced
  higher infection rates.
- **`date_value`** — positive and significant, capturing the **overall temporal
  growth** of the pandemic during the study period.
- **`lon`** — negative: as we move **east** (i.e. towards higher longitude
  in US), infection rates tend to be lower, i.e. rates are
  higher in the west.
- **Basis functions `B1`–`B9`** — capture residual spatial structure not
  explained by the linear trend surface; some are not individually
  significant here, but they matter jointly.

### Assessing Model Fit

Even though many coefficients are significant, the overall \(R^2\) is very low. This is a small figure and it reflects several issues:

- The response is a **count rate** and highly skewed — a
  Gaussian linear model is a poor match for its distribution.
- Only **one contextual factor** is included; many other socio-demographic
  drivers matter.

### In-Class Challenge

> **Challenge.** If you think that new cases grow and then drop during the period.  How can you modify the model?

---

# Poisson and Negative Binomial Regression for Count Data (Part 4)

## Why Not OLS for Counts?

The variable of concern in our regression is the number of new cases which is a **count** by itself. Counts have three properties that clash with the Gaussian linear model:

- They are **non-negative integers** — an OLS fit can (and will) predict
  negative values.
- Their distribution is typically **right-skewed** with many small values
  and a few very large ones.
- Their **variance grows with the mean** — larger counties with more cases
  also show more variability. OLS assumes constant variance.

An alternative of modelling is a **generalised linear model (GLM)** whose
response distribution is designed for counts. The two workhorses are the
**Poisson** and the **Negative Binomial** distributions.

## The Poisson Distribution

The Poisson distribution with mean \(\mu > 0\) has probability mass function

$$
P(Y = y) \;=\; \frac{\mu^{y}\, e^{-\mu}}{y!}, \qquad y = 0, 1, 2, \ldots
$$

Its defining property is the **equidispersion identity**:

$$
E[Y] \;=\; \mathrm{Var}[Y] \;=\; \mu.
$$

That is, once you know the mean, you know the variance. This is a very
strong (and often unrealistic) restriction.

## The Negative Binomial Distribution

The Negative Binomial (NB) distribution is another choice for count data.  We do not write down its probability mass function here but we highlight two main differences with Poisson.  Firstly the distribution has 2 parameters that gives more flexible shapes. Secondly, its variance is higher than mean that allows it to take care of **overdispersion** often found in count data.  For the second point, we have the following:

$$
E[Y] \;=\; \mu, \qquad
\mathrm{Var}[Y] \;=\; \mu + \frac{\mu^{2}}{\theta},
$$

where \(\theta > 0\) is a **dispersion parameter**. The extra term
\(\mu^{2}/\theta\) allows the variance to exceed the mean:

- As \(\theta \to \infty\), \(\mathrm{Var}[Y] \to \mu\) and NB collapses to
  Poisson.
- Small \(\theta\) implies large excess variance — heavy overdispersion.

## Numerical Illustration: Mean vs Variance

Let's simulate 10,000 draws from each distribution with the *same* mean
\(\mu = 5\) and compare their empirical variances.

```{r pois-nb-sim}
set.seed(1)

mu     <- 5
theta  <- 2       # NB dispersion parameter (smaller -> more dispersion)
n      <- 10000

y_pois <- rpois(n, lambda = mu)
y_nb   <- rnbinom(n, mu = mu, size = theta)

data.frame(
  Distribution = c("Poisson", "Negative Binomial"),
  Mean         = c(mean(y_pois),     mean(y_nb)),
  Variance     = c(var(y_pois),      var(y_nb)),
  "Var / Mean" = c(var(y_pois)/mean(y_pois), var(y_nb)/mean(y_nb)),
  check.names  = FALSE
)
```

Both samples have mean close to 5, but the NB sample has variance roughly
\(5 + 5^2/2 = 17.5\), i.e. about **3.5 times** the mean. The Poisson sample
has variance close to the mean, as it must.

## Visualising Both Distributions

Now compare the shapes of the two distributions side by side. We fix
\(\mu = 5\) and let \(\theta\) vary for the NB to see how dispersion
affects the shape.

```{r pois-nb-plot, fig.width=8, fig.height=4}
y_grid <- 0:25

pmf_pois   <- dpois(y_grid, lambda = mu)
pmf_nb_1   <- dnbinom(y_grid, mu = mu, size = 1)    # heavy dispersion
pmf_nb_2   <- dnbinom(y_grid, mu = mu, size = 2)    # moderate
pmf_nb_10  <- dnbinom(y_grid, mu = mu, size = 10)   # mild (close to Poisson)

op <- par(mfrow = c(1, 2))

# Left: Poisson pmf
plot(y_grid, pmf_pois, type = "h", lwd = 3, col = "steelblue",
     xlab = "y", ylab = "P(Y = y)",
     main = expression(paste("Poisson, ", mu == 5)))
points(y_grid, pmf_pois, pch = 16, col = "steelblue")

# Right: NB pmfs at different theta, plus Poisson for reference
matplot(y_grid,
        cbind(pmf_pois, pmf_nb_10, pmf_nb_2, pmf_nb_1),
        type = "b", pch = 16, lty = 1, lwd = 2,
        col = c("black", "darkgreen", "orange", "red"),
        xlab = "y", ylab = "P(Y = y)",
        main = expression(paste("NB (", mu == 5, "), varying ", theta)))
legend("topright", bty = "n", cex = 0.8,
       legend = c("Poisson", expression(theta == 10),
                  expression(theta == 2), expression(theta == 1)),
       col = c("black", "darkgreen", "orange", "red"),
       lty = 1, pch = 16)

par(op)
```

Notice:

- The Poisson pmf is fairly symmetric and tightly concentrated around
  \(\mu = 5\).
- As \(\theta\) decreases, the NB pmf becomes progressively **more skewed**,
  with a **higher probability of zero** and a **much heavier right tail**.
  These are exactly the features we typically see in real count data such as
  COVID-19 cases.
- For large \(\theta\), the NB curve is almost indistinguishable from the
  Poisson one — consistent with the theoretical limit
  \(\mathrm{Var}[Y] \to \mu\).

**Take-away.** If real data show more zeros and a heavier tail than
Poisson can produce, we need a distribution with a separate dispersion
parameter. That is exactly what NB gives us.

## Poisson Regression as a GLM

Poisson regression models the log of the mean as a linear function of
covariates:

$$
Y_i \sim \text{Poisson}(\mu_i), \qquad
\log(\mu_i) \;=\; \mathbf{x}_i^{\top} \boldsymbol{\beta}.
$$

The **log link** ensures \(\mu_i > 0\), and coefficients have a
multiplicative interpretation: a one-unit increase in \(x_k\) multiplies the
expected count by \(\exp(\beta_k)\).


### Fitting Poisson Regression to the COVID-19 Data

```{r poisson-fit}
covid19_reg_expand <- cbind(covid19_reg, new_cases=covid19_US$new_cases,
                            population=covid19_US$population)

m2_eq  <- new_cases ~ offset(log(population)) + lon + lat + week + pct_disab + B1 +
                         B2 + B3 + B4 + B5 + B6 +B7 +B8 +B9

m2 <- glm( m2_eq,data = covid19_reg_expand, 
                  family = poisson(link = "log"))

summary(m2)

```

Interestingly we find the latitude now becomes significant and all basis functions significant.  Is that a good fit?  We can have the following check for Poisson regression.

### Diagnosing Overdispersion

Under a correct Poisson model, the ratio of the residual deviance to the
residual degrees of freedom should be close to **1**. Any substantially
larger value is a red flag for overdispersion.

```{r poisson-dispersion}
m2$deviance / m2$df.residual
```

For the COVID-19 data this ratio is much greater than 1, confirming that
the variance is far exceeding its mean — exactly the situation
that the previous simulation warned us about.

## Negative Binomial Regression

NB regression uses the same log link as Poisson,

$$
\log(\mu_i) \;=\; \mathbf{x}_i^{\top} \boldsymbol{\beta},
$$

but assumes \(Y_i \sim \text{NB}(\mu_i, \theta)\), allowing
\(\mathrm{Var}[Y_i] = \mu_i + \mu_i^{2}/\theta\). The parameter \(\theta\) is
estimated from the data alongside \(\boldsymbol{\beta}\).

### Fitting NB Regression

```{r nb-fit}
library("MASS")

m3 <- glm.nb(m2_eq, data = covid19_reg_expand)
summary(m3)
```

The NB fit produces a **dramatic drop in residual deviance**
relative to Poisson which reflects a much better description of the extra variability
in the data.

