---
title: "CUSA1056 Space-Time Data Science"
subtitle: "Class 1: Visualization and Summaries of Spatio-Temporal Data"
author: "Dr. Remus Ho"
date: "`r Sys.Date()`"
output:
  pdf_document:
    latex_engine: xelatex
    toc: true
    toc_depth: 2
    number_sections: true
geometry: margin=1in
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE,
  fig.align = "center",
  fig.width = 6,      # safe PDF width
  fig.height = 4,
  out.width = "100%", # scale to page width
  fig.crop = FALSE
)
```

---

# Welcome & Overview

Welcome! This is the first class of CUSA1056. It is a challenging yet interesting course. I hope you will enjoy it and eventually come to appreciate the beauty of Statistics.

Our topic is spatio-temporal data. There are many examples:

- Daily temperatures recorded at multiple weather stations
- Monthly rainfall across Hong Kong
- Air quality readings in different districts over time
- Confirmed COVID-19 cases per city per week

There are special ways to describe and model these data, and we will learn them. However, these methods rely on an understanding of more fundamental statistical concepts. Therefore, from time to time, we will revisit core statistical topics. This will also help strengthen your foundational knowledge.

We will use R for statistical computation. We have chosen R because it offers rich libraries for a wide range of statistical procedures, especially for spatio-temporal studies. Don’t worry! We will keep the comments (descriptions after # in the codes) clear and will only expect you to understand specific commands, not to develop a thorough understanding of the entire programming language.

So without further ado, let’s get started!

**Today's plan:**

| Part | Topic |
|------|-------|
| 1 | **Visualization** — how to *see* patterns in space and time
| 2 | **Basic statistical concepts** — mean, variance, hypothesis testing, covariance, correlation
| 3 | **Numerical summaries** for spatio-temporal data

## Setup: installing and loading packages 

In the lab, R and RStudio (the IDE you will be using) are already installed on the computers. If you would like to use them at home, you can download R from  
https://cran.r-project.org/bin/windows/base/  
and RStudio from  
https://www.rstudio.com/products/rstudio/download/.  
Both are free of charge.

For R, you will need to install additional packages in order to use the methods and functions they provide. We have already installed the required packages on the lab computers. However, if you would like to practice at home, you can run the following commands **once** to install the packages we need:

```{r install, eval=FALSE}
install.packages(c("dplyr", "ggplot2", "maps", "patchwork", "tidyverse", "FRK", "lubridate", 
                   "MASS", "gstat", "sp", "spacetime", "animation"))

```

## Our dataset: NOAA daily maximum temperature

We will use the **NOAA data set**: daily maximum temperature (`Tmax`) in degrees Celsius (°C) at **152 weather stations** in the central and eastern USA, recorded in **2023**. We will focus on **May – September 2023** only in our examples.

```{r}
NOAA_df <- read.csv("NOAA_2023 v2.csv", header=TRUE, sep=",") # read in the dataset
# Create the new "date" column
NOAA_df$date <- as.Date(paste(NOAA_df$year, NOAA_df$month, NOAA_df$day, sep = "-"))
head(NOAA_df)
```


```{r}
library(dplyr) # need to apply functions in the package dplyr and have to load it here
Tmax <- filter(NOAA_df,               # only retrieve the variables and range of time we need
               var  == "Tmax",        # daily maximum temperature
               month %in% 5:9,        # May–September
               year  == 2023)         # year 2023

Tmax$t <- as.numeric(Tmax$date) - as.numeric(as.Date("2023-04-30"))          # day 1 = May 1, 2023

```

Each row is **one measurement**: one *station* on one *date*.

---

# PART 1 — Visualizing Spatio-Temporal Data

> "A picture is worth a thousand words."

Spatio-temporal data have **at least three dimensions** (longitude, latitude, time). We usually need several different plots to see the whole story.

## Spatial plots — snapshots in time

A **spatial plot** shows the value of a variable at every location, for **one time point**.

Let's plot daily maximum temperature on three different days: **May 1, June 1, July 1 of 2023**.

```{r spatial-plot, fig.height=4}
library(ggplot2) # a useful package for statistical plots
Tmax_1 <- subset(Tmax, t %in% c(1, 32, 62)) # retrieve data on the 3 days only

NOAA_plot <- ggplot(Tmax_1) +  # Create a ggplot object using Tmax_1 as the main dataset.
  geom_point(aes(x = lon, y = lat, colour = z), size = 2) +  # Add points with longitude and latitude, 
  # colored by z, with size 2.
  scale_colour_viridis_c(name = "degC", option = "turbo") +  # Apply a continuous viridis color scale 
  # labeled "degC" using the turbo palette.
  geom_path(data = map_data("state"),
            aes(x = long, y = lat, group = group)) +  # Overlay U.S. state boundaries using map data.
  facet_grid(~ date) +  # Create separate panels for each date.
  coord_fixed(xlim = c(-105, -75), ylim = c(25, 50)) +  # Fix the aspect ratio and limit the map to 
  #the specified longitude and latitude range.
  xlab("Longitude (deg)") + ylab("Latitude (deg)") +  # Label the x- and y-axes.
  theme_bw()  # Apply a black-and-white theme.

NOAA_plot  # Display the completed plot.

```

**What to look for:**

- **Color** shows the temperature (warmer = red, cooler = blue/green).
- **Trend in space:** temperatures are lower in the north and higher in the south.
- **Change over time:** the map changes from May 1 to July 1 as we get into summer.

> **Class discussion:** Where in the map do you see the biggest changes between May 1 and July 1?

## Time-series plots — one location, many time points

A **time-series plot** shows how a value changes over time, at **one location** (or several).

Let's pick 8 stations at random and look at their daily maximum temperatures throughout time:

```{r time-series, fig.height=4}
set.seed(12)
UIDs      <- unique(Tmax$id)  # Extract the unique station IDs from the Tmax dataset.
UIDs_sub  <- sample(UIDs, 8)  # Randomly select 8 station IDs.
Tmax_sub  <- filter(Tmax, id %in% UIDs_sub)  # Subset the data to include only 
# the selected station IDs.

TmaxTS <- ggplot(Tmax_sub) +  # Create a ggplot object using the subsetted data.
  geom_line(aes(x = t, y = z)) +  # Add a line plot of temperature (z) over time (t).
  facet_wrap(~ id, ncol = 4) +  # Create separate panels for each station ID arranged in 5 columns.
  xlab("Day number (from May 1, 2023)") +  # Label the x-axis.
  ylab("Tmax (degC)") +  # Label the y-axis.
  theme_bw() +  # Apply a black-and-white theme.
  theme(panel.spacing = unit(1, "lines"))  # Adjust spacing between panels.

TmaxTS  # Display the completed time series plot.
```

**What to look for:**

- All 8 stations basically show a **seasonal curve**: cooler in May and September, hotter around July.
- Some stations are **generally warmer** than others (their whole curve is higher). This reflects **latitude**: northern stations are cooler.
- The lines are **noisy** — that's day-to-day weather variability.

## Hovmöller plots — space vs. time in one picture

Sometimes we want to see *both* space and time in **one 2-D plot**. A **Hovmöller plot** does exactly that: one axis is space (say, latitude), the other axis is time, and color is the value.

Because our stations are irregular in space, we first bin them into latitude bands and average before plotting:

```{r hovmoller}
lim_lat  <- range(Tmax$lat)  # Compute the minimum and maximum latitude values.
lim_t    <- range(Tmax$t)  # Compute the minimum and maximum time values.
lat_axis <- seq(lim_lat[1], lim_lat[2], length = 25)  # Create 25 evenly spaced latitude grid points.
t_axis   <- seq(lim_t[1],  lim_t[2],  length = 100)  # Create 100 evenly spaced time grid points.

Tmax_grid       <- Tmax  # Make a copy of the original dataset.
dists           <- abs(outer(Tmax$lat, lat_axis, "-"))  
# Compute distances between each latitude and grid latitude.
Tmax_grid$lat   <- lat_axis[apply(dists, 1, which.min)]  
# Replace each latitude with its nearest grid latitude.

Tmax_lat_Hov <- group_by(Tmax_grid, lat, t) %>%  # Group data by grid latitude and time.
  summarise(z = mean(z, na.rm = TRUE), .groups = "drop")  
# Compute the mean temperature for each latitude–time pair.

Hovmoller_lat <- ggplot(Tmax_lat_Hov) +  # Create a ggplot object using the aggregated data.
  geom_tile(aes(x = lat, y = t, fill = z)) +  # Draw a heatmap with latitude on x, 
  # time on y, and color representing temperature.
  scale_fill_distiller(name = "degC", palette = "Spectral") +  
  # Apply a Spectral color palette labeled "degC".
  xlab("Latitude (deg)") + ylab("Day number (days)") +  # Label the axes.
  theme_bw()  # Apply a black-and-white theme.

Hovmoller_lat  # Display the Hovmöller diagram.
```

**How to read it:**

- The **horizontal axis** is latitude — moving right means moving north.
- The **vertical axis** is time — moving *up* means moving forward in time.
- **Color** is temperature.
- We can see the **north (right) is always cooler** than the south, and **temperature peaks in July and August**.

> **In-class exercise:** Can you produce a similar plot with latitude replaced by longitude?  What do you expect to see?

## Animations — a movie of your data

If we have many time points, we can string together spatial snapshots into a **movie**.

The R code below creates an HTML animation of daily maximum temperatures. It may take longer time (a few minutes) and you may run it after class:


```{r animation, fig.width=10, fig.height=7}
library(animation)  # Load the animation package for creating animated outputs.
library(dplyr)  # Load dplyr for data manipulation functions.
library(ggplot2)  # Load ggplot2 for plotting.
library(maps)  # Load maps for geographic boundary data.

# ---- Bounding box: from the data, with generous padding ----
lon_range <- range(Tmax$lon, na.rm = TRUE) + c(-3, 3)  # Compute longitude range and add padding.
lat_range <- range(Tmax$lat, na.rm = TRUE) + c(-2, 2)  # Compute latitude range and add padding.

# ---- Basemaps: countries + US state boundaries ----
world_map  <- map_data("world")  # Retrieve world map polygon data.
states_map <- map_data("state")  # Retrieve US state boundary data.

Tmax_t <- function(tau) {  # Define a function to plot temperature for a given time tau.
  Tmax_sub <- filter(Tmax, t == tau)  # Subset the data for the selected time.

  ggplot() +
    # country polygons (fills Canada / Mexico too so the frame isn't blank)
    geom_polygon(data = world_map,
                 aes(x = long, y = lat, group = group),
                 fill = "grey95", colour = "grey60", size = 0.3) +  # Draw filled country polygons.
    # US state boundaries on top of the country fill
    geom_path(data = states_map,
              aes(x = long, y = lat, group = group),
              colour = "grey55", size = 0.3) +  # Overlay US state boundaries.
    # temperature observations
    geom_point(data = Tmax_sub,
               aes(x = lon, y = lat, colour = z),
               size = 2.5, alpha = 0.9) +  # Plot temperature points colored by z.
    scale_colour_distiller(name = "z",
                           limits = c(-10, 50),
                           palette = "Spectral") +  # Apply a Spectral color scale with fixed limits.
    # Crop to the region and preserve geographic aspect ratio
    coord_quickmap(xlim = lon_range, ylim = lat_range, expand = TRUE) +  
    # Set map extent and aspect ratio.
    labs(title = paste("t =", tau),
         x = "Longitude", y = "Latitude") +  # Add title and axis labels.
    theme_bw(base_size = 13) +  # Apply a clean theme with base font size 13.
    theme(plot.margin = margin(5, 10, 5, 5))  # Adjust plot margins.
}

gen_anim <- function() {  # Define a function to generate animation frames.
  for (t in lim_t[1]:lim_t[2]) {  # Loop over all time values.
    plot(Tmax_t(t))  # Plot each time frame.
  }
}

# ---- Landscape frames sized for the US bounding box ----
ani.options(interval  = 0.2,
            ani.width  = 1000,   # Set animation frame width.
            ani.height = 700)    # Set animation frame height.

saveHTML(gen_anim(),
         autoplay = FALSE, loop = FALSE, verbose = FALSE,
         outdir = ".", htmlfile = "NOAA_anim.html")  # Save the animation as an HTML file.
```

Open `NOAA_anim.html` in your browser to watch the temperature map evolve day by day. You'll often see weather systems **drifting from west to east** — a moving pattern that a static plot cannot easily reveal.

Another example of animation: COVID 19 data animation from the a web page of John's Hopkins University (https://coronavirus.jhu.edu/data/animated-world-map )


### Part 1 summary

| Plot | What it shows |
|------|---------------|
| Spatial plot | The whole map at a **fixed time** |
| Time series | Value at a **fixed location** through time |
| Hovmöller | Space (1-D) **and** time together |
| Animation | Full spatial map **for every time** |

---

# PART 2 — Basic Statistical Concepts

Before we can go further, we need to understand a few  fundamental statistical ideas:

1. **Mean** — where is the "center"?
2. **Variance** — how spread out is the data?
3. **Hypothesis testing** - Is my guess right (supported by data)?
3. **Covariance** — do two variables move together?
4. **Correlation** — a standardized covariance (from −1 to +1)

All these ideas apply to **numbers**. In part 3, we will apply them to numbers *indexed by space and time*.

## The Mean (Average)

The **mean** is the "typical value" of a set of numbers.

If we have \(n\) numbers \(x_1, x_2, \dots, x_n\), then

$$
\bar{x} \;=\; \frac{1}{n}\sum_{i=1}^{n} x_i
\;=\; \frac{x_1 + x_2 + \cdots + x_n}{n}.
$$

**Example.** Suppose the temperature at your school on 5 days was 26, 28, 30, 27, 29 °C.

```{r mean-ex}
temps <- c(26, 28, 30, 27, 29)
mean(temps)
```

The mean is **28 °C** — the typical temperature.

## The Variance and Standard Deviation

The **variance** measures how much values *spread* around the mean.

$$
s^{2} \;=\; \frac{1}{n-1}\sum_{i=1}^{n}(x_i - \bar{x})^{2}.
$$

Steps to compute it:

1. Subtract the mean from each value (the "deviations").
2. Square them (so negatives don't cancel positives).
3. Add them all up.
4. Divide by \(n-1\).

The **standard deviation** is just \(s = \sqrt{s^{2}}\) and has the same units as the data.

```{r var-ex}
var(temps)        # variance
sd(temps)         # standard deviation
```

**Intuition:** A small variance means values are clustered near the mean; a big variance means they are spread out.

```{r var-picture, echo=FALSE, fig.height=3}
df <- data.frame(
  value = c(rnorm(50, 20, 1), rnorm(50, 20, 5)),
  group = rep(c("Small variance", "Large variance"), each = 50)
)
ggplot(df, aes(x = value, fill = group)) +
  geom_histogram(bins = 20, alpha = 0.7) +
  facet_wrap(~ group) +
  theme_bw() + xlab("value") + ylab("count") +
  theme(legend.position = "none")
```

## One-Sample t-Test for the Mean

We now know how to describe a sample with \(\bar{x}\) and \(s\). But a common scientific question is:

> "Is the *true* (population) mean equal to some value \(\mu_0\), or not?"

For example: *the historical July average temperature at this station was 27 °C. Is this year's July mean **different** from 27 °C, or is the difference just due to random day-to-day variation?*

The **one-sample t-test** answers exactly this question.

### The idea

We have a sample \(x_1, \dots, x_n\) with sample mean \(\bar{x}\) and sample SD \(s\). We compare \(\bar{x}\) against a hypothesized value \(\mu_0\).

- **Null hypothesis:** \(H_0:\ \mu = \mu_0\) (the true mean equals \(\mu_0\)).
- **Alternative:** \(H_1:\ \mu \neq \mu_0\) (two-sided), or \(\mu > \mu_0\) / \(\mu < \mu_0\) (one-sided).

### The test statistic

$$
t \;=\; \frac{\bar{x} - \mu_0}{s / \sqrt{n}}.
$$

The denominator \(s/\sqrt{n}\) is the **standard error of the mean** — how much \(\bar{x}\) itself is expected to wobble from sample to sample. So \(t\) is essentially asking:

> "How many standard errors is \(\bar{x}\) away from \(\mu_0\)?"

If \(H_0\) is true and the data are approximately normal, \(t\) follows a **Student's \(t\)-distribution with \(n-1\) degrees of freedom**.

### The p-value and decision rule

The **p-value** is the probability, *assuming \(H_0\) is true*, of getting a \(t\)-statistic at least as extreme as the one we observed.

- If **p-value \(< \alpha\)** (commonly \(\alpha = 0.05\)) \(\Rightarrow\) **reject** \(H_0\): the mean is significantly different from \(\mu_0\).
- If **p-value \(\ge \alpha\)** \(\Rightarrow\) **do not reject** \(H_0\): no strong evidence against \(\mu = \mu_0\).

### Example

Continuing with our 5 temperatures, test whether the true mean differs from 27 °C.

```{r ttest-ex}
temps <- c(26, 28, 30, 27, 29)

# Manual computation
xbar <- mean(temps)
s    <- sd(temps)
n    <- length(temps)
mu0  <- 27

t_stat <- (xbar - mu0) / (s / sqrt(n))
p_val  <- 2 * pt(-abs(t_stat), df = n - 1)   # two-sided

c(t = t_stat, df = n - 1, p_value = p_val)

# Built-in function
t.test(temps, mu = 27)
```

```{r ttest-plot, echo=FALSE, fig.height=3.2, fig.width=6}
library(ggplot2)

# Values from the example above
n       <- 5
df_t    <- n - 1
t_obs   <- (mean(temps) - 27) / (sd(temps) / sqrt(n))
alpha   <- 0.05
t_crit  <- qt(1 - alpha / 2, df = df_t)   # two-sided critical value

# t-density curve
xx   <- seq(-5, 5, length.out = 400)
dens <- data.frame(t = xx, d = dt(xx, df = df_t))

# Two-sided rejection region
rej_left  <- subset(dens, t <= -t_crit)
rej_right <- subset(dens, t >=  t_crit)

# Two-sided p-value region (|T| >= |t_obs|)
pv_left  <- subset(dens, t <= -abs(t_obs))
pv_right <- subset(dens, t >=  abs(t_obs))

ggplot(dens, aes(t, d)) +
  # rejection region (light red)
  geom_area(data = rej_left,  aes(t, d), fill = "red",    alpha = 0.25) +
  geom_area(data = rej_right, aes(t, d), fill = "red",    alpha = 0.25) +
  # p-value region (darker blue)
  geom_area(data = pv_left,   aes(t, d), fill = "steelblue", alpha = 0.55) +
  geom_area(data = pv_right,  aes(t, d), fill = "steelblue", alpha = 0.55) +
  # density curve
  geom_line(size = 0.8) +
  # observed t
  geom_vline(xintercept =  t_obs, colour = "steelblue", linetype = "dashed") +
  geom_vline(xintercept = -t_obs, colour = "steelblue", linetype = "dashed") +
  # critical values
  geom_vline(xintercept =  t_crit, colour = "red", linetype = "dotted") +
  geom_vline(xintercept = -t_crit, colour = "red", linetype = "dotted") +
  annotate("text", x = t_obs,  y = 0.02, label = sprintf("t = %.2f", t_obs),
           colour = "steelblue", hjust = -0.1, size = 3.5) +
  annotate("text", x = t_crit, y = 0.35,
           label = sprintf("t[crit] == %.2f", t_crit),
           parse = TRUE, colour = "red", hjust = -0.1, size = 3.5) +
  labs(
    title    = expression("Student's t-distribution with " * n - 1 == 4 * " df"),
    subtitle = "Red = rejection region (\u03b1 = 0.05, two-sided);  Blue = p-value area",
    x = "t", y = "density"
  ) +
  theme_bw()
```

**How to read this plot.**

- The **red dotted lines** at \(\pm t_{\text{crit}}\) mark the two-sided rejection region for \(\alpha = 0.05\) with \(n - 1 = 4\) degrees of freedom. If the observed \(t\) falls in the red tails, we reject \(H_0\).
- The **blue dashed lines** show our observed \(t\)-statistic (and its mirror image \(-t\)). The **blue shaded area** is the two-sided p-value: the probability under \(H_0\) of a \(t\) at least as extreme as ours.
- Here the blue lines lie **well inside** the red boundaries, and the blue area is large (~0.19). That is exactly what "we do not reject \(H_0\)" looks like visually.


**Reading the R output.** `t.test()` returns:

- the **t-statistic** and its **degrees of freedom**,
- the **p-value**,
- a **95% confidence interval** for the true mean,
- the sample mean \(\bar{x}\).

In this example \(\bar{x} = 28\), the p-value is around 0.19, so at \(\alpha = 0.05\) we **do not reject** \(H_0\): with only 5 days of data we cannot conclude the mean differs from 27 °C.

This is our first taste of **statistical inference**: going from a *sample* to a statement about the *population* it came from. Later we will need similar ideas of testing when we read the regression output for spatio-temporal data.

## Covariance: do two variables move together?

Now imagine we measure **two** things — for example, temperature (\(X\)) and ice-cream sales (\(Y\)) — on the same 5 days.

The **covariance** is

$$
\mathrm{cov}(X,Y) \;=\; \frac{1}{n-1}\sum_{i=1}^{n}(x_i - \bar{x})(y_i - \bar{y}).
$$

- If \(X\) is above its mean **whenever** \(Y\) is above *its* mean, covariance is **positive**.
- If \(X\) is above its mean **whenever** \(Y\) is below its mean, covariance is **negative**.
- If there is no pattern, covariance is near **zero**.


```{r cov-ex}
temp  <- c(26, 28, 30, 27, 29)          # °C
sales <- c(100, 120, 150, 110, 135)     # ice creams
cov(temp, sales)

```


The covariance is positive — as temperature rises, ice-cream sales rise too. Good news for the shop!

**Problem with covariance:** its size depends on the units. If we measured temperature in °F instead of °C, the covariance would change. So we need a **standardized** version.

## Correlation: covariance without units

The **Pearson correlation coefficient** is

$$
r \;=\; \frac{\mathrm{cov}(X,Y)}{s_X \, s_Y}.
$$

It is always between **−1 and +1**:

- \(r = +1\): perfect positive linear relationship
- \(r =  0\): no linear relationship
- \(r = -1\): perfect negative linear relationship

```{r cor-ex}
cor(temp, sales)
```

A correlation close to +1 confirms a strong positive linear relationship.

### Correlation gallery

```{r cor-gallery, echo=FALSE, fig.height=3.5}
set.seed(2)
n <- 100
x <- rnorm(n)
sim <- data.frame(
  x = rep(x, 4),
  y = c( x + rnorm(n,0,0.2),           # r ~ +1
        -x + rnorm(n,0,0.2),           # r ~ -1
         x + rnorm(n,0,1.5),           # r ~ +0.5
         rnorm(n))                     # r ~ 0
)
sim$r <- rep(c("r ~ +1", "r ~ -1", "r ~ +0.5", "r ~ 0"), each = n)
ggplot(sim, aes(x, y)) +
  geom_point(alpha = 0.5) + facet_wrap(~ r, nrow = 1) +
  theme_bw()
```

> **Warning!** Correlation only measures *linear* relationships. Two variables can be strongly related but have zero correlation (e.g., \(y = x^2\)).

### Part 2 summary

- **Mean** = center of the data.
- **Variance / SD** = spread around the mean.
- **One-sample t-test** = is the true mean equal to a specified value \(\mu_0\)?
- **Covariance** = do two variables vary together? (Units matter.)
- **Correlation** = same idea, standardized to \([-1, +1]\).

---

# PART 3 — Numerical Summaries for Spatio-Temporal Data

Now let's apply the ideas from Part 2 to our NOAA temperature data.

## Notation setup

We have observations \(Z(\mathbf{s}_i; t_j)\) at spatial locations \(\mathbf{s}_i,\ i=1,\dots,m\) and times \(t_j,\ j=1,\dots,T\).

- \(m\) = number of stations
- \(T\) = number of days

## Empirical Spatial Mean

For each **location**, average across **time**:

$$
\widehat{\mu}_{z,s}(\mathbf{s}_i)
 \;=\; \frac{1}{T}\sum_{j=1}^{T} Z(\mathbf{s}_i; t_j).
$$

This gives us **one number per station** — the typical temperature at that station over the summer.

```{r spat-mean}
spat_av <- group_by(Tmax, lat, lon) %>%  # Group the data by 
  # each spatial location (latitude and longitude).
  summarise(mu_emp = mean(z, na.rm = TRUE), .groups = "drop")  
# Compute the mean temperature at each location and drop grouping.

head(spat_av)  # Display the first few rows of the spatial averages dataset.
```

Now plot the spatial mean vs. latitude and vs. longitude:

```{r spat-mean-plot, fig.height=2, fig.width=4}
library(patchwork)  # Load patchwork to combine multiple ggplots.

lat_means <- ggplot(spat_av) +  # Create a scatter plot using the spatial averages data.
  geom_point(aes(lat, mu_emp)) +  # Plot mean temperature against latitude.
  xlab("Latitude (deg)") +  # Label the x-axis.
  ylab("Mean Tmax (degC)") + theme_bw()  # Label the y-axis and apply a clean theme.

lon_means <- ggplot(spat_av) +  # Create another scatter plot using the same data.
  geom_point(aes(lon, mu_emp)) +  # Plot mean temperature against longitude.
  xlab("Longitude (deg)") +  # Label the x-axis.
  ylab("Mean Tmax (degC)") + theme_bw()  # Label the y-axis and apply a clean theme.

lat_means + lon_means  # Display the two plots side by side.
```

**Observation:** the mean temperature **decreases clearly** with latitude (moving north = cooler). The decreasing trend with longitude is not so strong. So *latitude* is relatively a more useful "explanatory variable" for our data.

> **In-class exercise:** Can we study the minimum temperature similarly?  What do you see from the plots?

## Empirical Temporal Mean

For each **time**, average across **stations**:

$$
\widehat{\mu}_{z,t}(t_j)
 \;=\; \frac{1}{m}\sum_{i=1}^{m} Z(\mathbf{s}_i; t_j).
$$

This gives **one number per day** — the average temperature over the whole region on that day.

```{r time-mean}
Tmax_av <- group_by(Tmax, date) %>%  # Group the data by date.
  summarise(meanTmax = mean(z, na.rm = TRUE), .groups = "drop")  
# Compute the average maximum temperature for each date.

ggplot() +  # Initialize an empty ggplot object.
  geom_line(data = Tmax,
            aes(x = date, y = z, group = id),
            colour = "blue", alpha = 0.04) +  # Plot individual station time series with transparency.
  geom_line(data = Tmax_av,
            aes(x = date, y = meanTmax), linewidth = 0.9, group = 1) +  
  # Overlay the overall daily mean temperature as a thicker line.
  xlab("Date") + ylab("Maximum temperature (degC)") +  # Label the axes.
  theme_bw()  # Apply a clean black-and-white theme.
```

Each blue line is one station; the black line is the daily **regional average**. It clearly rises through spring/early summer, peaks in July/August, and falls in September — the seasonal cycle.

## Spatio-Temporal Covariogram and Semivariogram

### Motivation

For optimal prediction (kriging) that we shall learn later, we have to characterize the **joint spatio-temporal dependence structure** of a process. This dependence is described through:

- The **spatio-temporal covariogram**
- Its closely related counterpart, the **semivariogram**

Unlike ordinary covariance estimates, our goal is to describe dependence **as a function of spatial and temporal lags**.

- The temporal lag \( \tau \) is a scalar.
- The spatial lag \( \mathbf{h} \in \mathbb{R}^d \) is a vector representing displacement between two locations.

Under isotropy, spatial dependence may be expressed as a function of distance $h = \|\mathbf{h}\|$ only:

---

### Empirical Spatio-Temporal Covariogram

The empirical spatio-temporal covariogram for spatial lag \( \mathbf{h} \) and temporal lag \( \tau \) is

$$
\widehat{C}_Z(\mathbf{h}; \tau)
=
\frac{1}{|N_s(\mathbf{h})|}
\frac{1}{|N_t(\tau)|}
\sum_{s_i, s_k \in N_s(\mathbf{h})}
\sum_{t_j, t_\ell \in N_t(\tau)}
\bigl( Z(s_i; t_j) - \widehat{\mu}_{z,s}(s_i) \bigr)
\bigl( Z(s_k; t_\ell) - \widehat{\mu}_{z,s}(s_k) \bigr).
$$

where:

- \( N_s(\mathbf{h}) \): pairs of spatial locations within tolerance of \( \mathbf{h} \),
- \( N_t(\tau) \): pairs of time points within tolerance of \( \tau \),
- \( |N(\cdot)| \): number of elements in the set.

The covariogram measures **similarity**: large positive values indicate strong dependence at the given spatial and temporal lags.

---

### Empirical Spatio-Temporal Semivariogram

The **semivariogram** measures how *different* two observations are as a function of spatial and temporal separation.

$$
\widehat{\gamma}_z(\mathbf{h};\tau)
=
\frac{1}{2\,|N_s(\mathbf{h})|\,|N_t(\tau)|}
\sum_{s_i,s_k \in N_s(\mathbf{h})}
\sum_{t_j, t_\ell \in N_t(\tau)}
\bigl( Z(s_i;t_j) - Z(s_k;t_\ell) \bigr)^2.
$$

- Nearby locations and nearby times → small \( \widehat{\gamma}_z \).
- Far-apart locations or times → larger \( \widehat{\gamma}_z \).

The semivariogram measures **dissimilarity** rather than similarity.

---

### Relationship Between Covariogram and Semivariogram

Under second-order stationarity,

$$
\gamma(\mathbf{h};\tau)
=
C(\mathbf{0};0) - C(\mathbf{h};\tau).
$$

Thus,

- Large covariance \( \Rightarrow \) small semivariance.
- Small covariance \( \Rightarrow \) large semivariance.

When the variance exists, the two functions contain equivalent information.

---

### An example

To simplify computations, we use the June 2023 subset.

```{r}
library(sp); library(spacetime); library(gstat) 

TmaxJune <- subset(Tmax, month==6) %>%
  select(lat,lon,day,z)
TmaxJune <- na.omit(TmaxJune)

# --- Unique stations and full station × day grid --------------------------
stations <- unique(TmaxJune[, c("lon", "lat")])
rownames(stations) <- NULL
stations$id <- seq_len(nrow(stations))

days <- sort(unique(TmaxJune$day))

full <- expand.grid(id = stations$id, day = days)
full <- merge(full, stations, by = "id")
dat  <- merge(full, TmaxJune, by = c("lon", "lat", "day"), all.x = TRUE)
dat  <- dat[order(dat$day, dat$id), ]     # time-major order (required)
rownames(dat) <- NULL

# --- Spatial points (projected to km) -------------------------------------
sp_pts <- SpatialPoints(stations[, c("lon", "lat")],
                        CRS("+proj=longlat +datum=WGS84"))
sp_pts <- spTransform(sp_pts,
                      CRS("+proj=utm +zone=14 +datum=WGS84 +units=km"))

# --- Time index -----------------------------------------------------------
time_pts <- as.POSIXct(paste0("2023-06-", sprintf("%02d", days)), tz = "UTC")

# --- Build STFDF ----------------------------------------------------------
stfdf <- STFDF(sp   = sp_pts,
               time = time_pts,
               data = data.frame(z = dat$z, lat=dat$lat, lon=dat$lon, day=dat$day))

# --- Empirical semivariogram plot---------------------------------------------
vv <- variogram(z ~ lat + lon + day + I(day^2), data = stfdf,
                width  = 80,          # km
                cutoff = 1000,        # km
                tlags  = 0.01:6.01,   # days
                na.omit = TRUE)

plot(vv, main = "Semivariogram (spatial lag vs temporal lag)")

```


### Part 3 summary

| Quantity | Meaning |
|----------|---------|
| Empirical spatial mean | Average value at **each station** over time |
| Empirical temporal mean | Average value on **each day** across stations |
| Semivariogram | How different are values, as a function of space & time lag? |

---




