Below are codes for class 3 notes

```{r install, eval=FALSE}
install.packages(c("dplyr", "ggplot2", "maps", "patchwork", "tidyverse", "FRK", "lubridate", 
                   "MASS", "gstat", "sp", "spacetime", "animation"))

```

## Linear Regression

## Numerical Example: A Simple Simulated Dataset

To make the mechanics concrete, we simulate 40 points from the true
relationship

$$
y_i \;=\; 2 + 0.8\, x_i + \varepsilon_i,
\qquad \varepsilon_i \sim N(0, 1.2^{2}),
$$

fit an OLS line, and inspect the output.

```{r ols-example, fig.width=7, fig.height=4}
set.seed(2026)        # Set the random seed to 2026 to make the random number generation reproducible.

n <- 40               # Define the sample size n as 40 observations.

x <- runif(n, min = 0, max = 10)  
# Generate n random values for x from a uniform distribution between 0 and 10.

y <- 2 + 0.8 * x + rnorm(n, sd = 1.2)  
# Generate y values using a linear model with intercept 2 and slope 0.8 plus normally distributed 
# noise with standard deviation 1.2.

# Fit OLS
fit <- lm(y ~ x)      # Fit an ordinary least squares regression model of y on x.

summary(fit)          # Display a detailed summary of the fitted regression 
                      # model, including coefficients and statistics.
```

Inspecting the output, the estimated coefficients should be close to the
true values \(\beta_0 = 2\) and \(\beta_1 = 0.8\), with standard errors
that reflect the sample size and the noise level \(\sigma\).

### Visualising the Fit

```{r ols-plot, fig.width=7, fig.height=4}
plot(x, y, pch = 16, col = "steelblue",
     xlab = "x", ylab = "y",
     main = "OLS line with residuals")  
# Create a scatter plot of x and y with solid blue points, labeled axes, and a descriptive title.

abline(fit, col = "darkred", lwd = 2)  
# Add the fitted OLS regression line to the plot in dark red with a thicker line width.

# Draw residuals as vertical segments
y_hat <- fitted(fit)  
# Compute the fitted (predicted) values from the regression model and store them in y_hat.

segments(x, y, x, y_hat, col = "grey60", lty = 2)  
# Draw dashed grey vertical lines from each observed y value to its corresponding fitted value to
# #visualize residuals.

legend("topleft", bty = "n", cex = 0.9,
       legend = c("data", "OLS fit", "residuals"),
       col    = c("steelblue", "darkred", "grey60"),
       pch    = c(16, NA, NA),
       lty    = c(NA, 1, 2),
       lwd    = c(NA, 2, 1))  
# Add a legend in the top-left corner explaining the data points, OLS line, and residual lines with
# matching styles.
```

### Reconstructing OLS "by Hand"

To reinforce the matrix formulation, we can compute
\(\widehat{\boldsymbol{\beta}} = (\mathbf{X}^{\top}\mathbf{X})^{-1}\mathbf{X}^{\top}\mathbf{y}\)
directly and confirm we recover the same numbers as `lm()`.

```{r ols-by-hand}
X <- cbind(1, x)                              # design matrix
beta_hat <- solve(t(X) %*% X) %*% t(X) %*% y  # normal equations
beta_hat
coef(fit)                                     # compare with lm()
```

The two sets of numbers agree to machine precision. 

### A Non-Linear function Extension: Same Machinery, New Columns

```{r ols-quadratic, fig.width=7, fig.height=4}
set.seed(7)  
# Reproducibility

x  <- seq(0, 10, length.out = 60)  
# x grid (0–10, 60 pts)

y  <- 1 + 0.5 * x - 0.05 * x^2 + rnorm(length(x), sd = 0.6)  
# Quadratic model + noise

fit_quad <- lm(y ~ x + I(x^2))  
# Fit quadratic regression

coef(fit_quad)  
# Show coefficients

# Plot the fit
xg <- seq(0, 10, length.out = 200)  
# Fine grid for smooth curve

yg <- predict(fit_quad, newdata = data.frame(x = xg))  
# Predicted values

plot(x, y, pch = 16, col = "steelblue",
     xlab = "x", ylab = "y",
     main = "Quadratic fit via linear regression")  
# Scatter plot

lines(xg, yg, col = "darkred", lwd = 2)  
# Add fitted curve
```

# Deterministic Prediction — Inverse Distance Weighting (Part 2)

## IDW in R — using `gstat`

We will interpolate temperatures on a spatio-temporal grid, deliberately
**omitting 15 July 2023** from the training data, so that we can later see
how the method behaves on a day with no observations.

```{r idw-gstat}
library("dplyr")
library("ggplot2")
library("gstat")

NOAA_df <- read.csv("NOAA_2023 v2.csv", header = TRUE, sep = ",")  
# Load NOAA data from CSV

# Subset to Tmax, July 2023
Tmax <- filter(NOAA_df,
               var  == "Tmax" &
               month == 7     &
               year  == 2023)  
# Filter: Tmax, July 2023

# 20 x 20 spatial grid, 6 time points across the month
pred_grid <- expand.grid(lon = seq(-100, -80, length = 20),
                         lat = seq(  32,  46, length = 20),
                         day = seq(   5,  30, length = 6))  
# Create space–time prediction grid

# Remove 15 July from training data
Tmax_no_15 <- filter(Tmax, !(day == 15))  
# Exclude day 15 observations

# IDW with alpha = 4
Tmax_July_idw <- idw(formula   = z ~ 1,
                     locations = ~ lon + lat + day,
                     data      = Tmax_no_15,
                     newdata   = pred_grid,
                     idp       = 4)  
# Perform 3D IDW interpolation (power = 4)
```

We can then plot `Tmax_July_idw$var1.pred` on the prediction grid.
Note that on 15 July the predicted surface will look **smoother** than on
the surrounding days: with no data available on that day, the predictor
borrows information from adjacent days. This is a typical behaviour of any
predictor on days with no observations.

```{r}
library("dplyr")      # Data manipulation
library("ggplot2")    # Plotting
library("gstat")      # Spatial interpolation (IDW)
library("maps")       # Map outlines

# ---------------------------------------------------------------------------
# 1. Attach predictions to the grid
# ---------------------------------------------------------------------------

pred_grid1 <- pred_grid %>%
  mutate(z_pred = Tmax_July_idw$var1.pred)  
# Add IDW predictions to grid

pred_days <- sort(unique(pred_grid1$day))  
# Get the 6 prediction days

Tmax_days <- Tmax %>% 
  filter(day %in% pred_days)  
# Keep observations for those days

# ---------------------------------------------------------------------------
# 2. Basemap (state + country outlines)
# ---------------------------------------------------------------------------

states_map <- map_data("state")  
# Load US state boundaries

world_map  <- map_data("world")  
# Load world country boundaries

lon_range <- range(pred_grid1$lon)  
# Longitude limits

lat_range <- range(pred_grid1$lat)  
# Latitude limits

# ---------------------------------------------------------------------------
# 3. Faceted map: one panel per day
# ---------------------------------------------------------------------------

ggplot() +

  geom_raster(data = pred_grid1,
              aes(x = lon, y = lat, fill = z_pred)) +
  # IDW surface (filled grid)

  geom_polygon(data = world_map,
               aes(x = long, y = lat, group = group),
               fill = NA, colour = "black", size = 0.3) +
  # Country borders

  geom_path(data = states_map,
            aes(x = long, y = lat, group = group),
            colour = "grey40", size = 0.25) +
  # State borders

  geom_point(data = Tmax_days,
             aes(x = lon, y = lat, fill = z),
             shape = 21, colour = "white", size = 1.8, stroke = 0.3) +
  # Observed stations

  scale_fill_distiller(name = expression(T[max]~"("*degree*C*")"),
                       palette = "Spectral",
                       limits  = c(10, 45)) +
  # Color scale for temperature

  coord_quickmap(xlim = lon_range, ylim = lat_range, expand = FALSE) +
  # Fix map extent and aspect ratio

  facet_wrap(~ day, nrow = 2,
             labeller = labeller(day = function(d) paste("July", d, "2023"))) +
  # One panel per day

  labs(x = "Longitude", y = "Latitude",
       title = expression("IDW interpolation of "*T[max]*
                          " (July 2023), "*alpha*" = 4"),
       subtitle = "15 July was withheld from the training data") +
  # Titles and labels

  theme_bw(base_size = 12) +
  # Clean theme

  theme(strip.background = element_rect(fill = "grey90"),
        panel.grid.minor = element_blank())  
  # Minor styling adjustments
```

## Choosing \(\alpha\): Cross-Validation


### Tuning \(\alpha\) by LOOCV in R  

It takes very long.  Run this code at home.

```{r loocv-idw}
# Candidate values of alpha
alphas <- seq(2, 8, by = 1)
CV     <- numeric(length(alphas))

n <- nrow(Tmax)

for (a in seq_along(alphas)) {
  sq_err <- numeric(n)

  for (i in 1:n) {
    # leave observation i out
    train <- Tmax[-i, ]
    test  <- Tmax[ i, ]

    pred <- idw(formula   = z ~ 1,
                locations = ~ lon + lat + day,
                data      = train,
                newdata   = test,
                idp       = alphas[a],
                debug.level = 0)

    sq_err[i] <- (test$z - pred$var1.pred)^2
  }

  CV[a] <- mean(sq_err)
}

# LOOCV curve and optimum
plot(alphas, CV, type = "o",
     xlab = expression(alpha),
     ylab = expression(CV[(m)](alpha)))

alphas[which.min(CV)]   # optimal alpha
min(CV)                 # minimum CV score
```

# Linear Regression Models with Basis Functions (Part 3)

### Visualising the Two Bases (1D)

The following R snippet plots a degree-4 polynomial basis and a set of
Gaussian radial basis functions on the interval \([0, 1]\), to make the
global vs local distinction visible.

```{r basis-illustration, fig.width=8, fig.height=4}
op <- par(mfrow = c(1, 2))

s <- seq(0, 1, length = 400)

# --- Polynomial basis (global) ---
matplot(s, cbind(1, s, s^2, s^3, s^4), type = "l", lty = 1, lwd = 2,
        xlab = "s", ylab = expression(phi[k](s)),
        main = "Polynomial basis (global)")
legend("topleft",
       legend = c("1", "s", "s^2", "s^3", "s^4"),
       col = 1:5, lty = 1, bty = "n", cex = 0.8)

# --- Gaussian radial basis (local in practice) ---
centres <- seq(0.1, 0.9, by = 0.1)
theta   <- 0.08
Phi     <- sapply(centres, function(c) exp(-((s - c)^2) / theta^2))

matplot(s, Phi, type = "l", lty = 1, lwd = 2,
        xlab = "s", ylab = expression(phi[k](s)),
        main = "Gaussian radial basis (local)")

par(op)
```


## Application to a dataset of COVID cases in US

## Constructing Basis Functions with `FRK`

The **FRK** package can automatically build basis functions at multiple
resolutions.

```{r frk-basis, warning=FALSE}
library("FRK")
library("sp")

covid19_US <- read.csv("covid_US_weekly.csv", header=T, sep=",")
covid19_US$new_cases_per_ten_thousand <-covid19_US$new_cases/covid19_US$population*100000

# `covid19_US` is our spatio-temporal data frame with columns

# Build Gaussian radial basis functions at a single resolution
RB <- auto_basis(data = covid19_US[, c("lon", "lat")] %>% SpatialPoints(),
                nres = 1,
                type = "Gaussian")

# Evaluate basis functions at each observation location
RB_col <- eval_basis(basis = RB,
                s = covid19_US[, c("lon", "lat")] %>% as.matrix()) %>% as.matrix()

colnames(RB_col) <- paste0("B", 1:ncol(RB_col))
```

The columns of `RB_col` are simply new covariates (one per basis function) that
we can plug into a linear model.

## Application: COVID-19 Regression Across US counties

Our response variable is the weekly number of new cases per 100000 people in a county.  The independent variables are latitude, longitude, week number, basis function and one more covariate, the percentage of longterm illness in the county.

### Attach Basis Functions

```{r covid-basis}
covid19_reg <- cbind(covid19_US, RB_col) %>%
  dplyr::select(new_cases_per_ten_thousand,
                lon, lat, week, pct_disab,
                B1:B9)
```

### Fit the regression model

```{r covid-lm, warning=FALSE}
library(lubridate)

m1_eq  <- new_cases_per_ten_thousand ~ lon + lat + week + pct_disab + B1 + 
            B2 + B3 + B4 + B5 + B6 +B7 +B8 +B9

m1 <- lm(formula = m1_eq,
           data    = covid19_reg)

summary(m1)
```


### In-Class Challenge

> **Challenge.** If you think that new cases grow and then drop during the period.  How can you modify the model?  Try it below:

```{r}

```


---

# Poisson and Negative Binomial Regression for Count Data (Part 4)

## Numerical Illustration: Mean vs Variance

Let's simulate 10,000 draws from each distribution with the *same* mean
\(\mu = 5\) and compare their empirical variances.

```{r pois-nb-sim}
set.seed(1)

mu     <- 5
theta  <- 2       # NB dispersion parameter (smaller -> more dispersion)
n      <- 10000

y_pois <- rpois(n, lambda = mu)
y_nb   <- rnbinom(n, mu = mu, size = theta)

data.frame(
  Distribution = c("Poisson", "Negative Binomial"),
  Mean         = c(mean(y_pois),     mean(y_nb)),
  Variance     = c(var(y_pois),      var(y_nb)),
  "Var / Mean" = c(var(y_pois)/mean(y_pois), var(y_nb)/mean(y_nb)),
  check.names  = FALSE
)
```

Both samples have mean close to 5, but the NB sample has variance roughly
\(5 + 5^2/2 = 17.5\), i.e. about **3.5 times** the mean. The Poisson sample
has variance close to the mean, as it must.

## Visualising Both Distributions

Now compare the shapes of the two distributions side by side. We fix
\(\mu = 5\) and let \(\theta\) vary for the NB to see how dispersion
affects the shape.

```{r pois-nb-plot, fig.width=8, fig.height=4}
y_grid <- 0:25

pmf_pois   <- dpois(y_grid, lambda = mu)
pmf_nb_1   <- dnbinom(y_grid, mu = mu, size = 1)    # heavy dispersion
pmf_nb_2   <- dnbinom(y_grid, mu = mu, size = 2)    # moderate
pmf_nb_10  <- dnbinom(y_grid, mu = mu, size = 10)   # mild (close to Poisson)

op <- par(mfrow = c(1, 2))

# Left: Poisson pmf
plot(y_grid, pmf_pois, type = "h", lwd = 3, col = "steelblue",
     xlab = "y", ylab = "P(Y = y)",
     main = expression(paste("Poisson, ", mu == 5)))
points(y_grid, pmf_pois, pch = 16, col = "steelblue")

# Right: NB pmfs at different theta, plus Poisson for reference
matplot(y_grid,
        cbind(pmf_pois, pmf_nb_10, pmf_nb_2, pmf_nb_1),
        type = "b", pch = 16, lty = 1, lwd = 2,
        col = c("black", "darkgreen", "orange", "red"),
        xlab = "y", ylab = "P(Y = y)",
        main = expression(paste("NB (", mu == 5, "), varying ", theta)))
legend("topright", bty = "n", cex = 0.8,
       legend = c("Poisson", expression(theta == 10),
                  expression(theta == 2), expression(theta == 1)),
       col = c("black", "darkgreen", "orange", "red"),
       lty = 1, pch = 16)

par(op)
```

## Poisson Regression as a GLM

### Fitting Poisson Regression to the COVID-19 Data

```{r poisson-fit}
covid19_reg_expand <- cbind(covid19_reg, new_cases=covid19_US$new_cases,
                            population=covid19_US$population)

m2_eq  <- new_cases ~ offset(log(population)) + lon + lat + week + pct_disab + B1 +
                         B2 + B3 + B4 + B5 + B6 +B7 +B8 +B9

m2 <- glm( m2_eq,data = covid19_reg_expand, 
                  family = poisson(link = "log"))

summary(m2)

```


### Diagnosing Overdispersion

Under a correct Poisson model, the ratio of the residual deviance to the
residual degrees of freedom should be close to **1**. Any substantially
larger value is a red flag for overdispersion.

```{r poisson-dispersion}
m2$deviance / m2$df.residual
```

For the COVID-19 data this ratio is much greater than 1, confirming that
the variance is far exceeding its mean — exactly the situation
that the previous simulation warned us about.

## Negative Binomial Regression

### Fitting NB Regression

```{r nb-fit, warning=FALSE}
library("MASS")

m3 <- glm.nb(m2_eq, data = covid19_reg_expand)
summary(m3)
```


